# Multiple Linear Regression

## Introduction

Multiple linear regression extends the simple linear regression framework by allowing the response variable to depend on **several explanatory variables simultaneously**. This provides a flexible framework for describing more complex relationships and for constructing predictions using several sources of information.

In many applications, variation in a response cannot be adequately represented using a single explanatory variable. Economic growth, for example, may be related to investment, inflation, interest rates, and exports; academic performance may be related to study habits, socioeconomic characteristics, and prior preparation. Multiple linear regression provides a systematic way to represent these relationships within the same least-squares framework developed previously.

Importantly, the fundamental optimization problem does not change. Simple linear regression, polynomial regression, and multiple linear regression all use the same mathematical machinery. The main difference is the structure of the design matrix.

---

### The General Model

For observation $i$, the multiple linear regression model is

$$
y_i
=
\beta_0
+
\beta_1x_{1,i}
+
\beta_2x_{2,i}
+
\cdots
+
\beta_px_{p,i}
+
e_i,
\qquad
i=1,\ldots,n,
$$

where:

* $y_i$ is the $i$th observation of the **response variable**,
* $\beta_0$ is the **intercept**,
* $\beta_1,\ldots,\beta_p$ are the coefficients associated with the explanatory variables,
* $x_{1,i},\ldots,x_{p,i}$ are the values of the **explanatory variables** for observation $i$, and
* $e_i$ is the discrepancy between the observed response and the linear component of the model.

At this stage, no probabilistic assumptions are imposed on the errors. The model is treated as a deterministic representation of the observed response using a linear combination of the explanatory variables.

To establish notation that will be used throughout the chapter, define the matrix containing only the explanatory variables as

$$
\mathbf{X}_*
=
\begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_p
\end{bmatrix}
\in\mathbb{R}^{n\times p},
$$

where

$$
\mathbf{x}_j
=
\begin{bmatrix}
x_{j,1}\\
\vdots\\
x_{j,n}
\end{bmatrix}
\in\mathbb{R}^n,
\qquad
j=1,\ldots,p.
$$

The complete design matrix, including the column associated with the intercept, is

$$
\mathbf{X}
=
\begin{bmatrix}
\mathbf{1} & \mathbf{X}_*
\end{bmatrix}
\in\mathbb{R}^{n\times(p+1)}.
$$

Similarly, define

$$
\mathbf{y}
=
\begin{bmatrix}
y_1\\
\vdots\\
y_n
\end{bmatrix},
\qquad
\boldsymbol{\beta}
=
\begin{bmatrix}
\beta_0\\
\beta_1\\
\vdots\\
\beta_p
\end{bmatrix},
\qquad
\mathbf{e}
=
\begin{bmatrix}
e_1\\
\vdots\\
e_n
\end{bmatrix}.
$$

Then the model can be written compactly as

$$
\mathbf{y}
=
\mathbf{X}\boldsymbol{\beta}
+
\mathbf{e}.
$$

This is exactly the same matrix representation introduced previously for simple linear regression. The difference is that $\mathbf{X}$ now contains several explanatory-variable columns rather than only one.

---

### Interpretation and Objectives

Multiple linear regression has two important objectives in the present part of the course.

1. **Description** — to describe how the response varies with several explanatory variables considered simultaneously.

2. **Prediction** — to use several explanatory variables jointly to construct predictions of the response.

The coefficient $\beta_j$ describes how the linear component of the model changes with $x_j$ when the other explanatory variables are held fixed. In particular,

$$
\beta_0+\beta_1x_1+\cdots+\beta_jx_j+\cdots+\beta_px_p
$$

changes by $\beta_j$ when $x_j$ increases by one unit while the remaining explanatory variables remain unchanged.

This is often called a **partial** or **conditional interpretation** of the coefficient.

However, this interpretation is algebraic rather than automatically causal. Holding the remaining explanatory variables fixed allows us to isolate the contribution of one variable within the fitted linear model, but additional assumptions would be required to interpret that change as the causal effect of modifying the explanatory variable.

Thus, multiple regression allows us to separate the roles played by different columns of the design matrix without requiring a causal interpretation.

---

### Applications and Context

Multiple linear regression is useful whenever several quantities may help describe or predict a response.

For example, economic performance may be represented using explanatory variables such as inflation, unemployment, interest rates, government expenditure, and exports. Rather than fitting a separate regression for each explanatory variable, multiple regression allows all of them to enter the same linear model.

This generalization introduces several mathematical and computational issues that do not arise as clearly in simple regression.

One important issue is **dependence among the explanatory variables**. If two or more columns of $\mathbf{X}_*$ contain very similar information, it may become difficult to distinguish their individual contributions to the fitted model. In extreme cases, exact linear relationships among the columns can cause $\mathbf{X}$ to lose full column rank, so the usual least-squares solution is no longer unique.

Even when the columns are not exactly linearly dependent, strong relationships among them can affect the numerical stability and interpretation of the estimated coefficients. These issues will be studied through the linear algebra of the design matrix.

At this stage, our focus remains on the **least-squares optimization problem**: how the design matrix determines the fitted values, residuals, and coefficient estimates. Statistical inference and the probabilistic interpretation of these quantities will be introduced later.

---

### Connection to Polynomial Regression

We have already encountered a special case of multiple linear regression in **polynomial regression**.

For example, the quadratic model

$$
y_i
=
\beta_0
+
\beta_1x_i
+
\beta_2x_i^2
+
e_i
$$

can be written using the design matrix

$$
\mathbf{X}
=
\begin{bmatrix}
1 & x_1 & x_1^2\\
1 & x_2 & x_2^2\\
\vdots & \vdots & \vdots\\
1 & x_n & x_n^2
\end{bmatrix}.
$$

From the perspective of least squares, the columns containing $x$ and $x^2$ are treated in exactly the same way as two different explanatory-variable columns.

Polynomial regression therefore introduced the essential algebraic structure of multiple regression:

$$
\mathbf{y}
=
\mathbf{X}\boldsymbol{\beta}
+
\mathbf{e}.
$$

The difference is conceptual. In polynomial regression, the columns of $\mathbf{X}_*$ are transformations of a single original explanatory variable,

$$
x,\;x^2,\;\ldots,\;x^k,
$$

whereas in general multiple linear regression the columns

$$
\mathbf{x}_1,\ldots,\mathbf{x}_p
$$

may represent distinct explanatory variables.

Thus, multiple linear regression is the natural generalization of the framework already developed for polynomial regression. The least-squares machinery remains the same; what changes is the number and interpretation of the columns in the design matrix.



















## Example: Modeling GDP

As an applied example, consider the problem of describing and predicting a country’s Gross Domestic Product (GDP) using several economic indicators. We use the following explanatory variables:

* Inflation rate,
* Unemployment rate,
* Reference interest rate,
* Government spending as a percentage of GDP, and
* Exports as a percentage of GDP.

Unlike simple linear regression, where the relationship between the response and explanatory variable can be displayed in a single scatterplot, multiple regression involves several explanatory variables simultaneously. We therefore begin by examining the variables **pairwise**.

The following code reads the data and produces a scatterplot matrix containing a scatterplot for every pair of variables.

```{r paris-plot-gdp}
# Reads Data
dat <- read.csv(file = "Gdp Data.csv")

# Plot the scatterplots for each pair of variables
pairs(dat)
```

The scatterplot matrix provides an initial visual description of the dataset. In particular, it allows us to examine two types of relationships:

1. the pairwise relationship between GDP and each explanatory variable, and
2. the relationships among the explanatory variables themselves.

The second point becomes especially important in multiple regression. If two explanatory variables are strongly related to each other, then they may contain overlapping information about GDP. Consequently, their coefficients in a multiple regression may behave differently from the coefficients obtained when each variable is considered separately.

The same pairwise relationships can be summarized numerically using correlations.

```{r gdp-cor}
# Computes the correlation between variables
cor(dat)
```

The correlation matrix complements the scatterplot matrix by quantifying the strength and direction of the linear relationships between pairs of variables. Large positive or negative correlations between GDP and an explanatory variable suggest a strong pairwise linear relationship. Large correlations among explanatory variables indicate that some columns of the design matrix may contain similar information.

At this point, these quantities are used only as descriptive tools. The multiple-regression fit will determine how the explanatory variables behave when they are included in the model simultaneously.

---

### Simple Regressions

Before fitting the full multiple regression model, it is useful to fit a separate simple linear regression of GDP on each explanatory variable.

These models answer a different question from the multiple regression. A simple regression describes the relationship between GDP and one explanatory variable **without accounting for the remaining variables**.

For each explanatory variable, the code displays two plots:

* the scatterplot of GDP against the explanatory variable together with its fitted regression line, and
* the residuals from that simple regression.

The first plot helps us visualize the pairwise relationship. The second shows what remains unexplained after fitting that particular straight line.

#### Inflation Rate

```{r gdp-inf-fit}
# Fits with Inflation
outRegInf <- lm(gdp ~ inf, data = dat)
varVal    <- dat$inf
out       <- outRegInf
varNam    <- "Inflation Rate"
# Plots Regression Line and Scatterplot and residuals plot
par(mfrow = c(1, 2))
plot(x    = varVal,
     y    = dat$gdp,
     xlab = varNam,
     ylab = "GDP",
     pch  = 16)
abline(a   = out$coefficients[1],
       b   = out$coefficients[2],
       col = 'red',
       lwd = 2)
plot(x    = varVal,
     y    = out$residuals,
     xlab = varNam,
     ylab = "Residuals",
     pch  = 16)
abline(h   = 0,
       lwd = 2)
```

The fitted slope describes the pairwise linear relationship between inflation and GDP. It answers how the fitted value of GDP changes with inflation when inflation is the **only explanatory variable included in the model**.

The residual plot shows the differences between the observed GDP values and those predicted using inflation alone. Any remaining structure in the residuals indicates variation in GDP that this simple regression does not represent.

#### Unemployment Rate

```{r gdp-unp-fit}
# Fits with Unemployment
outRegUne <- lm(gdp ~ une, data = dat)
varVal    <- dat$une
out       <- outRegUne
varNam    <- "Unemployment Rate"
# Plots Regression Line and Scatterplot and residuals plot
par(mfrow = c(1, 2))
plot(x    = varVal,
     y    = dat$gdp,
     xlab = varNam,
     ylab = "GDP",
     pch  = 16)
abline(a   = out$coefficients[1],
       b   = out$coefficients[2],
       col = 'red',
       lwd = 2)
plot(x    = varVal,
     y    = out$residuals,
     xlab = varNam,
     ylab = "Residuals",
     pch  = 16)
abline(h   = 0,
       lwd = 2)
```

This regression describes the relationship between unemployment and GDP when unemployment is considered by itself.

As before, the fitted line summarizes the pairwise linear trend, while the residuals represent the component of GDP that is not captured by that one-variable linear fit.

#### Interest Rate

```{r gdp-int-fit}
# Fits with Interest Rate
outRegInt <- lm(gdp ~ int, data = dat)
varVal    <- dat$int
out       <- outRegInt
varNam    <- "Interest Rate"
# Plots Regression Line and Scatterplot and residuals plot
par(mfrow = c(1, 2))
plot(x    = varVal,
     y    = dat$gdp,
     xlab = varNam,
     ylab = "GDP",
     pch  = 16)
abline(a   = out$coefficients[1],
       b   = out$coefficients[2],
       col = 'red',
       lwd = 2)
plot(x    = varVal,
     y    = out$residuals,
     xlab = varNam,
     ylab = "Residuals",
     pch  = 16)
abline(h   = 0,
       lwd = 2)
```

The estimated slope now describes the pairwise relationship between the reference interest rate and GDP.

Importantly, this coefficient reflects both the relationship of the interest rate with GDP and any information that the interest rate shares with the other economic variables that have not yet been included in the model.

#### Government Spending

```{r gdp-gov-fit}
# Fits with Government Spending
outRegGov <- lm(gdp ~ gov, data = dat)
varVal    <- dat$gov
out       <- outRegGov
varNam    <- "Government Spending"
# Plots Regression Line and Scatterplot and residuals plot
par(mfrow = c(1, 2))
plot(x    = varVal,
     y    = dat$gdp,
     xlab = varNam,
     ylab = "GDP",
     pch  = 16)
abline(a   = out$coefficients[1],
       b   = out$coefficients[2],
       col = 'red',
       lwd = 2)
plot(x    = varVal,
     y    = out$residuals,
     xlab = varNam,
     ylab = "Residuals",
     pch  = 16)
abline(h   = 0,
       lwd = 2)
```

This fit examines GDP using government spending as the only explanatory variable.

As with the previous models, the estimated coefficient should be interpreted as a **simple-regression coefficient**. It does not yet describe the relationship between government spending and GDP after accounting for inflation, unemployment, interest rates, or exports.

#### Exports

```{r gdp-exp-fit}
# Fits with Exports
outRegExp <- lm(gdp ~ exp, data = dat)
varVal    <- dat$exp
out       <- outRegExp
varNam    <- "Exports"
# Plots Regression Line and Scatterplot and residuals plot
par(mfrow = c(1, 2))
plot(x    = varVal,
     y    = dat$gdp,
     xlab = varNam,
     ylab = "GDP",
     pch  = 16)
abline(a   = out$coefficients[1],
       b   = out$coefficients[2],
       col = 'red',
       lwd = 2)
plot(x    = varVal,
     y    = out$residuals,
     xlab = varNam,
     ylab = "Residuals",
     pch  = 16)
abline(h   = 0,
       lwd = 2)
```

The exports model provides the final pairwise regression. Its slope summarizes how the fitted GDP changes with exports when exports are considered alone.

Taken together, the five simple regressions provide useful benchmarks. They show what each explanatory variable appears to contribute when examined individually. We can now compare these results with a model in which all five explanatory variables are included simultaneously.

***
























### Multiple Regression

The simple regressions answer questions of the form

$$
\text{How is GDP related to } x_j \text{ when } x_j \text{ is used by itself?}
$$

Multiple regression answers a different question. It fits GDP using all explanatory variables simultaneously:

$$
\text{GDP}
=
\beta_0
+
\beta_1\text{Inflation}
+
\beta_2\text{Unemployment}
+
\beta_3\text{Interest}
+
\beta_4\text{Government Spending}
+
\beta_5\text{Exports}
+
e.
$$

Each coefficient now describes the change in the fitted linear component associated with one explanatory variable **while the values of the other explanatory variables are held fixed**.

We fit the full model and compare its coefficient estimates with those obtained from the corresponding simple regressions.

```{r gdp-all-ind-com}
outRegAll <- lm(gdp ~ inf + une + int + gov + exp, data = dat)

# Summary All
print("All Independent Variables")
summary(outRegAll)
print("Only Inflation Rate")
summary(outRegInf)
print("Only Unemployment Rate")
summary(outRegUne)
print("Only Interest Rate")
summary(outRegInt)
print("Only Government Spending")
summary(outRegGov)
print("Only Exports")
summary(outRegExp)


# Formats and Show Info Table
tab           <- cbind(outRegAll$coefficients[-1], c(outRegInf$coefficients[2],
                                                     outRegUne$coefficients[2],
                                                     outRegInt$coefficients[2],
                                                     outRegGov$coefficients[2],
                                                     outRegExp$coefficients[2]))
colnames(tab) <- c("All", "Individual")
rownames(tab) <- c("Inflation", "Unemployment", "Interest Rate", "Government Expenditure", "Exports")
knitr::kable(tab)
```

The final table places the multiple-regression coefficients next to the corresponding simple-regression coefficients. These coefficients need not be equal because they describe different fitted relationships.

For example, the simple-regression coefficient for inflation is obtained from a model containing only inflation,

$$
\text{GDP}
=
\beta_0+\beta_1\text{Inflation}+e,
$$

whereas the inflation coefficient in the multiple regression is obtained from a model in which unemployment, interest rates, government spending, and exports are also included.

Therefore, a change in the coefficient after additional explanatory variables are included is not itself surprising. It indicates that the explanatory variables contain overlapping information about the response.

A coefficient may become smaller, larger, or even change sign after the other explanatory variables are included. Such behavior motivates an important question that will be developed mathematically in this chapter:

> How does the relationship among the columns of the design matrix affect the estimated regression coefficients?

This is one of the central differences between simple and multiple regression. In simple regression, there is only one nonconstant explanatory-variable column. In multiple regression, the geometry of the entire collection

$$
\mathbf{x}_1,\ldots,\mathbf{x}_p
$$

matters.

The comparison therefore provides a first illustration of why multiple regression cannot be understood simply as several simple regressions performed at the same time. The coefficients are estimated **jointly**, and their values depend on the relationships among all columns of the design matrix.

***

### Summary

This introduction highlights the main ideas that distinguish multiple linear regression from simple linear regression:

1. **Explore the variables jointly.**
   Scatterplot matrices and correlations help identify pairwise relationships between the response and explanatory variables, as well as relationships among the explanatory variables themselves.

2. **Use simple regressions as benchmarks.**
   Fitting GDP separately on each explanatory variable shows how each variable behaves when considered alone.

3. **Fit all explanatory variables simultaneously.**
   Multiple regression estimates the coefficients jointly, so each coefficient describes the role of one explanatory variable while the others are held fixed within the linear model.

4. **Compare simple and multiple regression coefficients.**
   The coefficients can change substantially because the explanatory variables may contain overlapping information. Their values therefore depend on the relationships among all columns of the design matrix.

The central lesson is that multiple regression is **not simply a collection of simple regressions**. It is a single least-squares problem involving several explanatory variables at once.

This makes the structure of the design matrix particularly important. In the following sections, we study how the geometry and algebra of $\mathbf{X}$ determine the coefficient estimates, fitted values, residuals, and behavior of the multiple linear regression model.

***























## Least Squares Estimation

For least squares estimation, we solve the optimization problem

$$
Q(\bgb)
=
\sum_{i=1}^n
\left(
y_i-\beta_0-\sum_{j=1}^p\beta_jx_{j,i}
\right)^2
=
(\by-\bX\bgb)'(\by-\bX\bgb).
$$

The least-squares estimator is therefore defined as

$$
\hat{\bgb}
=
\underset{\bgb}{\operatorname{argmin}}
\;Q(\bgb).
$$

Because the problem has exactly the same matrix form as in simple linear regression, the solution is obtained in the same way.

Suppose that the full design matrix $\bX$ has full column rank,

$$
\operatorname{rank}(\bX)=p+1.
$$

Then $\bX'\bX$ is invertible, and the unique least-squares solution is

$$
\hat{\bgb}
=
(\bX'\bX)^{-1}\bX'\by.
$$

The formula is therefore unchanged from the matrix formulation of simple linear regression. The difference is only the dimension of the coefficient vector. In multiple linear regression,

$$
\bgb
=
\left(
\beta_0,
\beta_1,
\beta_2,
\ldots,
\beta_p
\right)',
$$

and consequently,

$$
\hat{\bgb}
=
\left(
\hat{\beta}_0,
\hat{\beta}_1,
\hat{\beta}_2,
\ldots,
\hat{\beta}_p
\right)'.
$$

Thus,

$$
\boxed{
\text{same least-squares optimization}
\quad\Longrightarrow\quad
\text{same matrix solution}
}
$$

with more explanatory variables. This is why working in matrix form is particularly useful.




















## Properties of the Estimates

As with simple linear regression, we distinguish between the **fitted values** and the **residuals**.

Given the least-squares estimator

$$
\hat{\bgb}
=
(\bX'\bX)^{-1}\bX'\by,
$$

the fitted values are

$$
\hat{\by}
=
\bX\hat{\bgb},
$$

and the residuals are

$$
\hat{\be}
=
\by-\hat{\by}
=
\by-\bX\hat{\bgb}.
$$

We can also express the fitted values as

$$
\hat{\by}
=
\bX(\bX'\bX)^{-1}\bX'\by
=
\bH\by,
$$

where

$$
\bH
=
\bX(\bX'\bX)^{-1}\bX'
$$

is called the **hat matrix**, because it transforms $\by$ into $\hat{\by}$. It is also called the **projection matrix**, since it projects $\by$ onto the column space of $\bX$.

The properties of the least-squares solution follow primarily from one important consequence of the normal equations:

$$
\boxed{
\hat{\bgb}
=
(\bX'\bX)^{-1}\bX'\by
\quad\Longrightarrow\quad
\bX'\hat{\be}=\bzero.
}
$$

When the design matrix includes an intercept, this identity implies that the residuals sum to zero, are orthogonal to every explanatory-variable vector and to the fitted values, and that the mean of the fitted values equals the mean of the observed responses.

```{proposition, prp-mlr-properties-estimates, name="Properties of the Least-Squares Estimates"}
Suppose that the design matrix $\bX$ has full column rank and includes a column of ones for the intercept. Then the least-squares solution satisfies the following properties:

1. Each component of $\hat{\bgb}$ is a linear combination of the entries of $\by$.

2. The sum of the residuals is equal to zero,

   $$
   \sum_{i=1}^n\hat e_i=0.
   $$

3. The residual vector is orthogonal to every explanatory-variable vector,

   $$
   \hat{\be}\perp\bx_j,
   \qquad j=1,\ldots,p.
   $$

4. The residual vector is orthogonal to the fitted-value vector,

   $$
   \hat{\be}\perp\hat{\by}.
   $$

5. The mean of the fitted values equals the mean of the observed responses,

   $$
   \bar y=\overline{\hat y},
   $$

   where

   $$
   \overline{\hat y}
   =
   \frac{1}{n}\sum_{i=1}^n\hat y_i.
   $$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-properties-estimates}

**1. Linearity of the coefficient estimates**

To show that each component of $\hat{\bgb}$ is a linear combination of the entries of $\by$, we need to express the estimator as

$$
\hat{\bgb}=\bA\by
$$

for some matrix $\bA$ that does not depend on $\by$.

Let

$$
\bA
=
(\bX'\bX)^{-1}\bX'.
$$

Then,

$$
\hat{\bgb}
=
(\bX'\bX)^{-1}\bX'\by
=
\bA\by.
$$

When the design matrix $\bX$ is held fixed, $\bA$ is also fixed. Therefore, each component of $\hat{\bgb}$ is a linear combination of the entries of $\by$.

**2. The normal equations and the sum of the residuals**

The remaining properties follow from the normal equations.

Starting with the least-squares solution,

\begin{align*}
\hat{\bgb}
  &= (\bX'\bX)^{-1}\bX'\by
    & \text{least-squares solution} \\
\bX'\bX\hat{\bgb}
  &= \bX'\by
    & \text{multiply by } \bX'\bX \\
\bX'\by-\bX'\bX\hat{\bgb}
  &= \bzero
    & \text{rearrange} \\
\bX'(\by-\bX\hat{\bgb})
  &= \bzero
    & \text{factor } \bX' \\
\bX'\hat{\be}
  &= \bzero
    & \text{def. residuals}.
\end{align*}

This is the central identity from which the remaining properties follow.

Recall that the design matrix includes the column of ones,

$$
\bX
=
\begin{bmatrix}
\bones & \bx_1 & \cdots & \bx_p
\end{bmatrix}.
$$

Consequently,

$$
\bX'
=
\begin{bmatrix}
\bones'\\
\bx_1'\\
\vdots\\
\bx_p'
\end{bmatrix}.
$$

Expanding the product $\bX'\hat{\be}$, we obtain

$$
\bX'\hat{\be}
=
\begin{bmatrix}
\bones'\hat{\be}\\
\bx_1'\hat{\be}\\
\vdots\\
\bx_p'\hat{\be}
\end{bmatrix}
=
\begin{bmatrix}
0\\
0\\
\vdots\\
0
\end{bmatrix}.
$$

The first component of this equality gives

$$
\bones'\hat{\be}=0.
$$

Since $\bones$ is the vector of ones,

$$
\bones'\hat{\be}
=
\sum_{i=1}^n\hat e_i.
$$

Therefore,

$$
\boxed{
\sum_{i=1}^n\hat e_i=0.
}
$$

This establishes the second property.

**3. Orthogonality of the residuals to the explanatory variables**

Using the same expansion of the normal equations,

$$
\bX'\hat{\be}
=
\begin{bmatrix}
\bones'\hat{\be}\\
\bx_1'\hat{\be}\\
\vdots\\
\bx_p'\hat{\be}
\end{bmatrix}
=
\bzero,
$$

we see that its remaining components give

$$
\bx_j'\hat{\be}=0,
\qquad j=1,\ldots,p.
$$

Recall that two vectors are orthogonal when their inner product is zero. Therefore,

$$
\boxed{
\hat{\be}\perp\bx_j,
\qquad j=1,\ldots,p.
}
$$

This establishes the third property.

**4. Orthogonality of the residuals to the fitted values**

To show that the residuals are orthogonal to the fitted values, we need to establish that

$$
\hat{\be}'\hat{\by}=0.
$$

Recall that

$$
\hat{\by}
=
\bX\hat{\bgb}.
$$

Therefore,

\begin{align*}
\hat{\be}'\hat{\by}
  &= \hat{\be}'\bX\hat{\bgb}
    & \text{def. fitted values} \\
  &= (\bX'\hat{\be})'\hat{\bgb}
    & \text{transpose rule} \\
  &= \bzero'\hat{\bgb}
    & \text{normal equations} \\
  &= 0.
\end{align*}

Consequently,

$$
\boxed{
\hat{\be}\perp\hat{\by}.
}
$$

This establishes the fourth property.

Notice that the fitted values are linear combinations of the columns of $\bX$. Since the residuals are orthogonal to every column of $\bX$, they must also be orthogonal to the fitted-value vector.

**5. Equality of the means**

Finally, we show that

$$
\bar y=\overline{\hat y}.
$$

Recall from the second property that

$$
\bones'\hat{\be}=0.
$$

Using the definition of the residuals,

\begin{align*}
\bones'\hat{\be}
  &= \bones'(\by-\hat{\by})
    & \text{def. residuals} \\
  &= \bones'\by-\bones'\hat{\by}
    & \text{distributivity} \\
  &= \sum_{i=1}^n y_i-\sum_{i=1}^n\hat y_i
    & \text{def. } \bones \\
  &= n\bar y-n\overline{\hat y}
    & \text{def. means}.
\end{align*}

Since $\bones'\hat{\be}=0$, we obtain

\begin{align*}
n\bar y-n\overline{\hat y}
  &=0\\
n\bar y
  &=n\overline{\hat y}\\
\bar y
  &=\overline{\hat y}.
\end{align*}

Therefore,

$$
\boxed{
\bar y=\overline{\hat y}.
}
$$

This establishes the fifth property.

Notice that the second and fifth properties depend on the presence of the intercept in the model. Including the intercept ensures that $\bones$ is a column of $\bX$, and therefore the normal equations imply

$$
\bones'\hat{\be}=0.
$$

Without an intercept, the residuals are still orthogonal to the columns of $\bX$, but they are not necessarily orthogonal to $\bones$. Consequently, their sum need not equal zero, and the mean of the fitted values need not equal the mean of the observed responses.
```

</details>

The main result of this subsection is therefore the hierarchy

$$
\boxed{
\hat{\bgb}
=
(\bX'\bX)^{-1}\bX'\by
\quad\Longrightarrow\quad
\bX'\hat{\be}=\bzero.
}
$$

Since the design matrix includes the intercept and all explanatory-variable vectors, the normal equations imply

$$
\boxed{
\bX'\hat{\be}=\bzero
\quad\Longrightarrow\quad
\begin{cases}
\displaystyle\sum_{i=1}^n\hat e_i=0,\\[6pt]
\hat{\be}\perp\bx_j,\quad j=1,\ldots,p,\\[4pt]
\hat{\be}\perp\hat{\by},\\[4pt]
\bar y=\overline{\hat y}.
\end{cases}
}
$$

Thus, the properties of the residuals and fitted values are direct consequences of a single matrix identity.

























## Multiple $R^2$

As with simple linear regression, we can evaluate the fit of a multiple linear regression model by decomposing the total variability of the response into two components: the variability accounted for by the fitted model and the variability remaining in the residuals.

Throughout this section, we assume that the full design matrix $\mathbf{X}$ includes a column of ones for the intercept and has full column rank.

We begin by defining three sums of squares.

```{definition, def-mlr-total-sum-of-squares, name="Total Sum of Squares"}
The **Total Sum of Squares**, $SS_{tot}$, measures the total variability of the observed responses around their sample mean:

$$
SS_{tot}
=
(\mathbf{y}-\bar y\mathbf{1})'
(\mathbf{y}-\bar y\mathbf{1})
=
\sum_{i=1}^n(y_i-\bar y)^2.
$$
```

```{definition, def-mlr-residual-sum-of-squares, name="Residual Sum of Squares"}
The **Residual Sum of Squares**, $SS_{res}$, measures the variability remaining after fitting the regression model:

$$
SS_{res}
=
(\mathbf{y}-\hat{\mathbf{y}})'
(\mathbf{y}-\hat{\mathbf{y}})
=
\sum_{i=1}^n(y_i-\hat y_i)^2.
$$
```

```{definition, def-mlr-explained-sum-of-squares, name="Explained Sum of Squares"}
The **Explained Sum of Squares**, $SS_{reg}$, measures the variability of the fitted values around the mean of the observed responses:

$$
SS_{reg}
=
(\hat{\mathbf{y}}-\bar y\mathbf{1})'
(\hat{\mathbf{y}}-\bar y\mathbf{1})
=
\sum_{i=1}^n(\hat y_i-\bar y)^2.
$$
```

The properties of the least-squares estimates established in the previous section allow us to decompose the total sum of squares into the explained and residual sums of squares.

```{proposition, prp-mlr-sum-of-squares, name="Sum-of-Squares Decomposition"}
Suppose that the multiple linear regression model includes an intercept. Then the least-squares fitted values satisfy

$$
SS_{tot}=SS_{reg}+SS_{res}.
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-sum-of-squares}
We begin with the definition of the total sum of squares. Adding and subtracting the fitted-value vector gives

\begin{align*}
SS_{tot}
  &=
  (\mathbf{y}-\bar y\mathbf{1})'
  (\mathbf{y}-\bar y\mathbf{1})
  \\
  &=
  (\mathbf{y}-\hat{\mathbf{y}}
  +\hat{\mathbf{y}}-\bar y\mathbf{1})'
  (\mathbf{y}-\hat{\mathbf{y}}
  +\hat{\mathbf{y}}-\bar y\mathbf{1})
  \\
  &=
  (\mathbf{y}-\hat{\mathbf{y}})'
  (\mathbf{y}-\hat{\mathbf{y}})
  \\
  &\quad+
  (\mathbf{y}-\hat{\mathbf{y}})'
  (\hat{\mathbf{y}}-\bar y\mathbf{1})
  \\
  &\quad+
  (\hat{\mathbf{y}}-\bar y\mathbf{1})'
  (\mathbf{y}-\hat{\mathbf{y}})
  \\
  &\quad+
  (\hat{\mathbf{y}}-\bar y\mathbf{1})'
  (\hat{\mathbf{y}}-\bar y\mathbf{1}).
\end{align*}

Recall from the properties of the least-squares estimates that

$$
\hat{\mathbf{e}}'\hat{\mathbf{y}}=0
$$

and, because the model includes an intercept,

$$
\hat{\mathbf{e}}'\mathbf{1}=0.
$$

Therefore, the first cross-product is

\begin{align*}
(\mathbf{y}-\hat{\mathbf{y}})'
(\hat{\mathbf{y}}-\bar y\mathbf{1})
  &=
  \hat{\mathbf{e}}'
  (\hat{\mathbf{y}}-\bar y\mathbf{1})
  \\
  &=
  \hat{\mathbf{e}}'\hat{\mathbf{y}}
  -
  \bar y\hat{\mathbf{e}}'\mathbf{1}
  \\
  &=
  0.
\end{align*}

The other cross-product is its transpose and is also zero. Consequently,

\begin{align*}
SS_{tot}
  &=
  (\mathbf{y}-\hat{\mathbf{y}})'
  (\mathbf{y}-\hat{\mathbf{y}})
  \\
  &\quad+
  (\hat{\mathbf{y}}-\bar y\mathbf{1})'
  (\hat{\mathbf{y}}-\bar y\mathbf{1})
  \\
  &=
  SS_{res}+SS_{reg}.
\end{align*}

Thus,

$$
\boxed{
SS_{tot}=SS_{reg}+SS_{res}.
}
$$
```

</details>


This decomposition allows us to define the coefficient of determination for multiple linear regression.


```{definition, def-mlr-multiple-r-squared, name="Multiple Coefficient of Determination"}
Suppose that the multiple linear regression model includes an intercept and that $SS_{tot}>0$.

The **Multiple Coefficient of Determination**, $R^2$, is defined as the proportion of the total variability of the observed responses accounted for by the fitted regression model:

$$
R^2
=
\frac{SS_{reg}}{SS_{tot}}.
$$

Using the sum-of-squares decomposition,

$$
SS_{tot}=SS_{reg}+SS_{res},
$$

we obtain the equivalent expression

$$
R^2
=
1-\frac{SS_{res}}{SS_{tot}}.
$$
```

The multiple $R^2$ is an in-sample measure of fit. It describes the reduction in squared error achieved by the full regression model relative to the intercept-only model. It does not, by itself, establish predictive performance on new observations.

Using the sum-of-squares decomposition, we also obtain

\begin{align*}
1
  &=
  \frac{SS_{reg}+SS_{res}}{SS_{tot}}
  \\
  &=
  R^2+\frac{SS_{res}}{SS_{tot}}.
\end{align*}

Therefore,

$$
\boxed{
R^2
=
1-\frac{SS_{res}}{SS_{tot}}.
}
$$

The multiple $R^2$ describes the proportion of the observed variability around $\bar y$ accounted for by the fitted linear model. It is an in-sample measure of fit and does not, by itself, establish causal relationships or predictive performance on new observations.

***

### The Intercept-Only Model

Before expressing $R^2$ using projection matrices, it is useful to introduce a reference model.

Consider a regression model containing only an intercept,

$$
y_i=\beta_0+e_i,
\qquad i=1,\ldots,n.
$$

Its design matrix is

$$
\mathbf{X}_0=\mathbf{1},
$$

and its least-squares problem is

$$
\min_{\beta_0}
(\mathbf{y}-\mathbf{1}\beta_0)'
(\mathbf{y}-\mathbf{1}\beta_0).
$$

The least-squares solution is

$$
\hat\beta_0
=
(\mathbf{1}'\mathbf{1})^{-1}
\mathbf{1}'\mathbf{y}
=
\bar y.
$$

Consequently, the fitted values from the intercept-only model are

$$
\hat{\mathbf{y}}_0
=
\bar y\mathbf{1}.
$$

This model assigns the same predicted response, $\bar y$, to every observation. Its residual sum of squares is precisely

$$
SS_{tot}
=
(\mathbf{y}-\bar y\mathbf{1})'
(\mathbf{y}-\bar y\mathbf{1}).
$$

Thus, the total sum of squares can be interpreted as the minimum squared error obtained when we predict the response using only a constant.

Adding explanatory variables enlarges the set of possible fitted-value vectors. This observation will allow us to establish both the range of $R^2$ and its behavior when additional variables are included.

***

### Sums of Squares in Matrix Form

Recall that the hat matrix for the full multiple linear regression model is

$$
\mathbf{H}
=
\mathbf{X}
(\mathbf{X}'\mathbf{X})^{-1}
\mathbf{X}',
$$

so that

$$
\hat{\mathbf{y}}=\mathbf{H}\mathbf{y}.
$$

Similarly, the hat matrix for the intercept-only model is

$$
\mathbf{H}_0
=
\mathbf{1}
(\mathbf{1}'\mathbf{1})^{-1}
\mathbf{1}'
=
\frac{1}{n}\mathbf{1}\mathbf{1}'.
$$

It follows that

$$
\mathbf{H}_0\mathbf{y}
=
\bar y\mathbf{1}.
$$

The matrices $\mathbf{H}$ and $\mathbf{H}_0$ project the response vector onto the column spaces of the full and intercept-only design matrices, respectively.

We can now express all three sums of squares using these projection matrices.

```{proposition, prp-mlr-projection-sums-of-squares, name="Sums of Squares Using Projection Matrices"}
Let

$$
\mathbf{H}
=
\mathbf{X}
(\mathbf{X}'\mathbf{X})^{-1}
\mathbf{X}'
$$

be the hat matrix of the full multiple linear regression model, where $\mathbf{X}$ has full column rank and includes an intercept.

Let

$$
\mathbf{H}_0
=
\frac{1}{n}\mathbf{1}\mathbf{1}'
$$

be the hat matrix of the intercept-only model.

Then the three sums of squares can be written as

$$
\boxed{
SS_{tot}
=
\mathbf{y}'
(\mathbf{I}-\mathbf{H}_0)
\mathbf{y},
}
$$

$$
\boxed{
SS_{res}
=
\mathbf{y}'
(\mathbf{I}-\mathbf{H})
\mathbf{y},
}
$$

and

$$
\boxed{
SS_{reg}
=
\mathbf{y}'
(\mathbf{H}-\mathbf{H}_0)
\mathbf{y}.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-projection-sums-of-squares}
We first establish the necessary properties of the projection matrices.

Recall that

$$
\mathbf{H}
=
\mathbf{X}
(\mathbf{X}'\mathbf{X})^{-1}
\mathbf{X}'.
$$

Since $\mathbf{X}'\mathbf{X}$ is symmetric and invertible, its inverse is also symmetric. Therefore,

$$
\mathbf{H}'=\mathbf{H}.
$$

Furthermore,

\begin{align*}
\mathbf{H}^2
  &=
  \mathbf{X}
  (\mathbf{X}'\mathbf{X})^{-1}
  \mathbf{X}'
  \mathbf{X}
  (\mathbf{X}'\mathbf{X})^{-1}
  \mathbf{X}'
  \\
  &=
  \mathbf{X}
  (\mathbf{X}'\mathbf{X})^{-1}
  (\mathbf{X}'\mathbf{X})
  (\mathbf{X}'\mathbf{X})^{-1}
  \mathbf{X}'
  \\
  &=
  \mathbf{X}
  (\mathbf{X}'\mathbf{X})^{-1}
  \mathbf{X}'
  \\
  &=
  \mathbf{H}.
\end{align*}

Thus, $\mathbf{H}$ is symmetric and idempotent.

Consequently, $\mathbf{I}-\mathbf{H}$ is also symmetric, and

\begin{align*}
(\mathbf{I}-\mathbf{H})^2
  &=
  \mathbf{I}-2\mathbf{H}+\mathbf{H}^2
  \\
  &=
  \mathbf{I}-\mathbf{H}.
\end{align*}

Therefore, $\mathbf{I}-\mathbf{H}$ is idempotent.

**Residual Sum of Squares**

The residual vector can be written as

\begin{align*}
\hat{\mathbf{e}}
  &=
  \mathbf{y}-\hat{\mathbf{y}}
  \\
  &=
  \mathbf{y}-\mathbf{H}\mathbf{y}
  \\
  &=
  (\mathbf{I}-\mathbf{H})\mathbf{y}.
\end{align*}

Hence,

\begin{align*}
SS_{res}
  &=
  \hat{\mathbf{e}}'\hat{\mathbf{e}}
  \\
  &=
  \bigl[(\mathbf{I}-\mathbf{H})\mathbf{y}\bigr]'
  \bigl[(\mathbf{I}-\mathbf{H})\mathbf{y}\bigr]
  \\
  &=
  \mathbf{y}'
  (\mathbf{I}-\mathbf{H})'
  (\mathbf{I}-\mathbf{H})
  \mathbf{y}
  \\
  &=
  \mathbf{y}'
  (\mathbf{I}-\mathbf{H})^2
  \mathbf{y}
  \\
  &=
  \mathbf{y}'
  (\mathbf{I}-\mathbf{H})
  \mathbf{y}.
\end{align*}

Therefore,

$$
\boxed{
SS_{res}
=
\mathbf{y}'
(\mathbf{I}-\mathbf{H})
\mathbf{y}.
}
$$

**Total Sum of Squares**

For the intercept-only model, we have

$$
\mathbf{H}_0
=
\frac{1}{n}\mathbf{1}\mathbf{1}'.
$$

This matrix is symmetric. It is also idempotent, since

\begin{align*}
\mathbf{H}_0^2
  &=
  \frac{1}{n^2}
  \mathbf{1}\mathbf{1}'
  \mathbf{1}\mathbf{1}'
  \\
  &=
  \frac{1}{n^2}
  \mathbf{1}
  (\mathbf{1}'\mathbf{1})
  \mathbf{1}'
  \\
  &=
  \frac{1}{n}
  \mathbf{1}\mathbf{1}'
  \\
  &=
  \mathbf{H}_0.
\end{align*}

Consequently, $\mathbf{I}-\mathbf{H}_0$ is symmetric and idempotent:

$$
(\mathbf{I}-\mathbf{H}_0)^2
=
\mathbf{I}-2\mathbf{H}_0+\mathbf{H}_0^2
=
\mathbf{I}-\mathbf{H}_0.
$$

Recall that

$$
\mathbf{H}_0\mathbf{y}
=
\bar y\mathbf{1}.
$$

Therefore,

\begin{align*}
SS_{tot}
  &=
  (\mathbf{y}-\bar y\mathbf{1})'
  (\mathbf{y}-\bar y\mathbf{1})
  \\
  &=
  \bigl[(\mathbf{I}-\mathbf{H}_0)\mathbf{y}\bigr]'
  \bigl[(\mathbf{I}-\mathbf{H}_0)\mathbf{y}\bigr]
  \\
  &=
  \mathbf{y}'
  (\mathbf{I}-\mathbf{H}_0)'
  (\mathbf{I}-\mathbf{H}_0)
  \mathbf{y}
  \\
  &=
  \mathbf{y}'
  (\mathbf{I}-\mathbf{H}_0)
  \mathbf{y}.
\end{align*}

Thus,

$$
\boxed{
SS_{tot}
=
\mathbf{y}'
(\mathbf{I}-\mathbf{H}_0)
\mathbf{y}.
}
$$

**Explained Sum of Squares**

Finally, using the sum-of-squares decomposition,

\begin{align*}
SS_{reg}
  &=
  SS_{tot}-SS_{res}
  \\
  &=
  \mathbf{y}'
  (\mathbf{I}-\mathbf{H}_0)
  \mathbf{y}
  -
  \mathbf{y}'
  (\mathbf{I}-\mathbf{H})
  \mathbf{y}
  \\
  &=
  \mathbf{y}'
  \bigl[
    (\mathbf{I}-\mathbf{H}_0)
    -
    (\mathbf{I}-\mathbf{H})
  \bigr]
  \mathbf{y}
  \\
  &=
  \mathbf{y}'
  (\mathbf{H}-\mathbf{H}_0)
  \mathbf{y}.
\end{align*}

Therefore,

$$
\boxed{
SS_{reg}
=
\mathbf{y}'
(\mathbf{H}-\mathbf{H}_0)
\mathbf{y}.
}
$$
```

</details>

The three sums of squares therefore have a unified matrix representation:

| Sum of squares | Matrix expression |
|---|---|
| Total | $SS_{tot}=\mathbf{y}'(\mathbf{I}-\mathbf{H}_0)\mathbf{y}$ |
| Residual | $SS_{res}=\mathbf{y}'(\mathbf{I}-\mathbf{H})\mathbf{y}$ |
| Explained | $SS_{reg}=\mathbf{y}'(\mathbf{H}-\mathbf{H}_0)\mathbf{y}$ |

Using these identities, the multiple $R^2$ can be written as

$$
\boxed{
R^2
=
1-
\frac{
\mathbf{y}'(\mathbf{I}-\mathbf{H})\mathbf{y}
}{
\mathbf{y}'(\mathbf{I}-\mathbf{H}_0)\mathbf{y}
}.
}
$$

This expression compares the squared error from the full regression model with the squared error from the intercept-only model.

***

### Range of Multiple $R^2$

The optimization interpretation of the two models allows us to establish the possible values of $R^2$.

```{proposition, prp-mlr-r2-range, name="Range of Multiple R-Squared"}
Suppose that the multiple linear regression model includes an intercept and that

$$
SS_{tot}>0.
$$

Then the coefficient of determination satisfies

$$
\boxed{
0\leq R^2\leq 1.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-r2-range}
The residual sum of squares from the full regression model is the minimum value of its least-squares objective:

$$
SS_{res}
=
\min_{\boldsymbol{\beta}}
(\mathbf{y}-\mathbf{X}\boldsymbol{\beta})'
(\mathbf{y}-\mathbf{X}\boldsymbol{\beta}).
$$

Similarly, the total sum of squares is the minimum squared error from the intercept-only model:

$$
SS_{tot}
=
\min_{\beta_0}
(\mathbf{y}-\mathbf{1}\beta_0)'
(\mathbf{y}-\mathbf{1}\beta_0).
$$

Because the full model includes an intercept, every fitted-value vector available to the intercept-only model is also available to the full model.

In particular, the full model can reproduce any constant fit by setting all the explanatory-variable coefficients equal to zero.

Therefore, minimizing over the larger set of possible fitted values cannot produce a larger minimum:

$$
SS_{res}\leq SS_{tot}.
$$

Both sums of squares are nonnegative, so

$$
0\leq SS_{res}\leq SS_{tot}.
$$

Since $SS_{tot}>0$, dividing by $SS_{tot}$ gives

$$
0
\leq
\frac{SS_{res}}{SS_{tot}}
\leq
1.
$$

Recall that

$$
R^2
=
1-\frac{SS_{res}}{SS_{tot}}.
$$

Consequently,

$$
\boxed{
0\leq R^2\leq 1.
}
$$
```

</details>

The condition $SS_{tot}>0$ means that the observed responses are not all identical. When all responses are equal, the usual definition of $R^2$ has a zero denominator and does not apply.

The extreme values have simple interpretations. If $R^2=0$, the full model does not reduce the residual sum of squares relative to the intercept-only model. If $R^2=1$, the full model fits every observed response exactly, so $SS_{res}=0$.

***

### Adding an Explanatory Variable

One consequence of the least-squares formulation is that adding an explanatory variable cannot increase the minimum residual sum of squares when both models are fitted to the same observations.

The additional variable expands the set of possible fitted-value vectors. At worst, the expanded model can reproduce the original fit by assigning a coefficient of zero to the new variable.

```{proposition, prp-mlr-additional-variable-r2, name="Effect of Adding an Explanatory Variable on R-Squared"}
Consider a multiple linear regression model with an intercept and design matrix $\mathbf{X}$. Let its residual sum of squares and coefficient of determination be $SS_{res}$ and $R^2$, respectively.

Now consider an expanded model obtained by adding another explanatory-variable vector $\mathbf{z}$:

$$
\mathbf{X}_+
=
\begin{bmatrix}
\mathbf{X} & \mathbf{z}
\end{bmatrix}.
$$

Let $SS_{res,+}$ and $R_+^2$ denote the corresponding quantities for the expanded model, fitted using the same observations.

Then

$$
\boxed{
SS_{res,+}\leq SS_{res}.
}
$$

Furthermore, if $SS_{tot}>0$, then

$$
\boxed{
R_+^2\geq R^2.
}
$$

Thus, adding an explanatory variable cannot decrease the ordinary in-sample $R^2$.
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-additional-variable-r2}
For the original regression model, the residual sum of squares is

$$
SS_{res}
=
\min_{\boldsymbol{\beta}}
\left\|
\mathbf{y}-\mathbf{X}\boldsymbol{\beta}
\right\|^2.
$$

For the expanded model, introduce the coefficient $\gamma$ associated with the additional explanatory variable $\mathbf{z}$.

Its residual sum of squares is

$$
SS_{res,+}
=
\min_{\boldsymbol{\beta},\gamma}
\left\|
\mathbf{y}
-
\mathbf{X}\boldsymbol{\beta}
-
\gamma\mathbf{z}
\right\|^2.
$$

Notice that the expanded optimization problem includes every candidate fitted-value vector from the original problem. We can always recover the original model by choosing

$$
\gamma=0.
$$

Therefore,

\begin{align*}
SS_{res,+}
  &=
  \min_{\boldsymbol{\beta},\gamma}
  \left\|
  \mathbf{y}
  -
  \mathbf{X}\boldsymbol{\beta}
  -
  \gamma\mathbf{z}
  \right\|^2
  \\
  &\leq
  \min_{\boldsymbol{\beta}}
  \left\|
  \mathbf{y}
  -
  \mathbf{X}\boldsymbol{\beta}
  -
  0\mathbf{z}
  \right\|^2
  \\
  &=
  \min_{\boldsymbol{\beta}}
  \left\|
  \mathbf{y}
  -
  \mathbf{X}\boldsymbol{\beta}
  \right\|^2
  \\
  &=
  SS_{res}.
\end{align*}

Thus,

$$
\boxed{
SS_{res,+}\leq SS_{res}.
}
$$

Since both regressions use the same response observations, their total sum of squares is the same.

Provided that $SS_{tot}>0$,

\begin{align*}
R_+^2
  &=
  1-\frac{SS_{res,+}}{SS_{tot}}
  \\
  &\geq
  1-\frac{SS_{res}}{SS_{tot}}
  \\
  &=
  R^2.
\end{align*}

Therefore,

$$
\boxed{
R_+^2\geq R^2.
}
$$
```

</details>


The inequality need not be strict. For example, if the new explanatory-variable vector is already a linear combination of the columns of $\mathbf{X}$, the expanded model has exactly the same set of possible fitted-value vectors. In this case, adding the variable does not change the residual sum of squares or $R^2$, although the expanded design matrix no longer has full column rank if the original matrix was already full rank.

More generally, equality occurs whenever the additional variable fails to reduce the minimum residual sum of squares.

This result reveals an important limitation of ordinary $R^2$ as a model-comparison criterion. **Adding explanatory variables can never decrease the in-sample $R^2$**, even when the additional variables do not improve predictions for new observations.

Consequently, a larger $R^2$ does not necessarily imply better predictive performance.

***

### Connection to the Geometric Interpretation

The two derivations in this section express the same decomposition from different perspectives.

The first uses the orthogonality of the residual vector to the fitted values and the intercept column. The second uses the projection matrices associated with the full and intercept-only models.

Because the column space of the intercept-only model is contained in the column space of the full model, the matrices also satisfy

$$
\mathbf{H}\mathbf{H}_0
=
\mathbf{H}_0\mathbf{H}
=
\mathbf{H}_0.
$$

Consequently,

$$
(\mathbf{H}-\mathbf{H}_0)^2
=
\mathbf{H}-\mathbf{H}_0.
$$

Thus, $\mathbf{H}-\mathbf{H}_0$ is itself a symmetric, idempotent matrix. In particular,

$$
\hat{\mathbf{y}}-\bar y\mathbf{1}
=
(\mathbf{H}-\mathbf{H}_0)\mathbf{y}.
$$

The decomposition

$$
\mathbf{y}-\bar y\mathbf{1}
=
(\hat{\mathbf{y}}-\bar y\mathbf{1})
+
\hat{\mathbf{e}}
$$

therefore separates the total centered response vector into two orthogonal components.

Their squared lengths are $SS_{reg}$ and $SS_{res}$, and the squared length of their sum is $SS_{tot}$.

This provides the connection to the geometric interpretation of multiple regression developed in the following section.


### Uncentered $R^2$: Regression Without an Intercept

The usual definition of $R^2$ compares the residual sum of squares of a regression model with that of the intercept-only model,

$$
y_i=\beta_0+e_i.
$$

When the full model includes an intercept, the intercept-only model is a special case: we can recover it by setting all explanatory-variable coefficients equal to zero. Consequently, the full model cannot have a larger minimum residual sum of squares.

However, consider a regression model **without an intercept**,

$$
\mathbf{y}
=
\mathbf{X}\boldsymbol{\beta}+\mathbf{e},
$$

where $\mathbf{X}$ contains only explanatory-variable columns. The intercept-only model is no longer necessarily a special case of this model, since $\mathbf{1}$ need not belong to $\mathcal{C}(\mathbf{X})$.

Therefore, the residual sum of squares of the regression model may exceed $SS_{tot}$, and the usual centered $R^2$ may be negative.

Instead, we can compare the regression model with the simplest possible prediction: predicting zero for every observation. This reference model has no parameters to estimate, so its optimization problem is trivial:

$$
\min_{\mathbf{v}\in\{\mathbf{0}\}}
\|\mathbf{y}-\mathbf{v}\|^2
=
\|\mathbf{y}\|^2
=
\sum_{i=1}^n y_i^2.
$$

Unlike the intercept-only model, the zero prediction is always available to a regression model without an intercept by setting $\boldsymbol{\beta}=\mathbf{0}$.

```{definition, def-mlr-uncentered-r-squared, name="Uncentered Coefficient of Determination"}
Suppose that the regression model does not include an intercept and that $\mathbf{y}'\mathbf{y}>0$.

The **Uncentered Coefficient of Determination** is defined as

$$
R_{\mathrm{unc}}^2
=
1-
\frac{SS_{res}}{\mathbf{y}'\mathbf{y}},
$$

where

$$
SS_{res}
=
\min_{\boldsymbol{\beta}}
\|\mathbf{y}-\mathbf{X}\boldsymbol{\beta}\|^2.
$$

Thus, uncentered $R^2$ measures the proportional reduction in squared error relative to predicting zero for every observation.
```

The use of the zero prediction as a reference allows us to establish the range of uncentered $R^2$ without requiring an intercept.

```{proposition, prp-mlr-uncentered-r-squared-range, name="Range of Uncentered R-Squared"}
Suppose that $\mathbf{y}'\mathbf{y}>0$. Then the uncentered coefficient of determination satisfies

$$
\boxed{
0\leq R_{\mathrm{unc}}^2\leq 1.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-uncentered-r-squared-range}
The residual sum of squares is the minimum value of the least-squares objective,

$$
SS_{res}
=
\min_{\boldsymbol{\beta}}
\|\mathbf{y}-\mathbf{X}\boldsymbol{\beta}\|^2.
$$

Notice that $\boldsymbol{\beta}=\mathbf{0}$ is always a feasible choice, which produces the zero prediction. Therefore,

\begin{align*}
SS_{res}
  &=
  \min_{\boldsymbol{\beta}}
  \|\mathbf{y}-\mathbf{X}\boldsymbol{\beta}\|^2
  \\
  &\leq
  \|\mathbf{y}-\mathbf{X}\mathbf{0}\|^2
  \\
  &=
  \mathbf{y}'\mathbf{y}.
\end{align*}

Since squared norms are nonnegative,

$$
0\leq SS_{res}\leq\mathbf{y}'\mathbf{y}.
$$

Dividing by $\mathbf{y}'\mathbf{y}>0$, we obtain

$$
0
\leq
\frac{SS_{res}}{\mathbf{y}'\mathbf{y}}
\leq
1.
$$

Consequently,

$$
\boxed{
0\leq R_{\mathrm{unc}}^2\leq 1.
}
$$
```

</details>


**Geometric interpretation.** The least-squares residuals remain orthogonal to the fitted values even without an intercept. Therefore, the Pythagorean theorem gives

$$
\|\mathbf{y}\|^2
=
\|\hat{\mathbf{y}}\|^2
+
\|\hat{\mathbf{e}}\|^2.
$$

Hence, the uncentered coefficient of determination can also be expressed as

$$
\boxed{
R_{\mathrm{unc}}^2
=
\frac{\hat{\mathbf{y}}'\hat{\mathbf{y}}}
{\mathbf{y}'\mathbf{y}}.
}
$$

This differs from centered $R^2$, which measures improvement relative to predicting the sample mean. Consequently, centered and uncentered $R^2$ should not be compared directly when evaluating different models.

As with ordinary $R^2$, adding explanatory variables cannot decrease uncentered $R^2$ when the same observations are used, since the additional variables enlarge the set of possible fitted values. This does not guarantee better predictions for new observations.

***

### Adjusted $R^2$

We established that adding an explanatory variable cannot increase the residual sum of squares. Consequently, ordinary $R^2$ cannot decrease when explanatory variables are added, even if the additional variables provide very little improvement in fit.

This motivates a modification of $R^2$ that accounts for the number of coefficients in the regression model.

Consider a multiple linear regression model with $p$ explanatory variables and an intercept. Assume that its design matrix $\mathbf{X}$ has full column rank and that $n>p+1$.

The column space of $\mathbf{X}$ has dimension $p+1$. Consequently, the residuals belong to its orthogonal complement, which has dimension

$$
n-(p+1)=n-p-1.
$$

The residual sum of squares therefore has $n-p-1$ **residual degrees of freedom**. Similarly, the total sum of squares is based on deviations from the sample mean and has $n-1$ degrees of freedom.

Instead of comparing the original sums of squares, we can compare them after dividing by their respective degrees of freedom.

```{definition, def-mlr-adjusted-r-squared, name="Adjusted Coefficient of Determination"}
Suppose that a multiple linear regression model includes an intercept, has $p$ explanatory variables, and satisfies $n>p+1$ and $SS_{tot}>0$.

The **Adjusted Coefficient of Determination** is defined as

$$
\boxed{
R_{\mathrm{adj}}^2
=
1-
\frac{SS_{res}/(n-p-1)}
{SS_{tot}/(n-1)}.
}
$$

Equivalently, using the ordinary coefficient of determination,

$$
\boxed{
R_{\mathrm{adj}}^2
=
1-
\frac{n-1}{n-p-1}(1-R^2).
}
$$
```

Unlike ordinary $R^2$, adjusted $R^2$ accounts for the number of explanatory variables through the residual degrees of freedom. A new variable must provide a sufficient reduction in residual sum of squares to offset the loss of one degree of freedom.

Since

$$
\frac{n-1}{n-p-1}\geq 1,
$$

we also have

$$
R_{\mathrm{adj}}^2\leq R^2\leq 1.
$$

Adjusted $R^2$ can be negative, particularly when the model provides little improvement over the intercept-only model relative to the number of explanatory variables included.

***

#### Adding an Explanatory Variable

We can establish precisely when adding another explanatory variable improves adjusted $R^2$.

```{proposition, prp-mlr-additional-variable-adjusted-r2, name="Effect of Adding an Explanatory Variable on Adjusted R-Squared"}
Consider a regression model with an intercept, $p$ explanatory variables, and residual sum of squares $SS_{res}$.

Suppose that adding one explanatory variable produces a full-column-rank model with residual sum of squares $SS_{res,+}$. Assume that both models use the same observations, $SS_{tot}>0$, and $n>p+2$.

Then adjusted $R^2$ increases if and only if

$$
\boxed{
SS_{res}-SS_{res,+}
>
\frac{SS_{res}}{n-p-1}.
}
$$

Thus, the reduction in residual sum of squares must exceed the original residual sum of squares divided by its degrees of freedom.
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-additional-variable-adjusted-r2}
The original model has $n-p-1$ residual degrees of freedom, while the expanded model has $n-p-2$.

Because both models use the same response observations, their total sum of squares is identical. Therefore,

\begin{align*}
R_{\mathrm{adj},+}^2>R_{\mathrm{adj}}^2
&\iff
1-\frac{SS_{res,+}/(n-p-2)}{SS_{tot}/(n-1)}
>
1-\frac{SS_{res}/(n-p-1)}{SS_{tot}/(n-1)}
\\
&\iff
\frac{SS_{res,+}}{n-p-2}
<
\frac{SS_{res}}{n-p-1}
\\
&\iff
(n-p-1)SS_{res,+}
<
(n-p-2)SS_{res}
\\
&\iff
SS_{res}-SS_{res,+}
>
\frac{SS_{res}}{n-p-1}.
\end{align*}

This establishes the result.
```

</details>

Ordinary $R^2$ increases whenever an additional variable reduces the residual sum of squares. Adjusted $R^2$, in contrast, increases only when that reduction exceeds the threshold established above.

**Adjusted $R^2$ is therefore useful for comparing models with different numbers of explanatory variables**, provided they are fitted to the same response observations. However, it remains an in-sample measure of fit. A higher adjusted $R^2$ does not guarantee better predictions on new observations; cross-validation provides a more direct way to evaluate predictive performance.


***


### Summary

The coefficient of determination, $R^2$, measures the reduction in squared error achieved by a regression model relative to a reference prediction. The appropriate reference depends on whether the model includes an intercept.

**1. Regression with an intercept**

The reference is the intercept-only model, which predicts $\bar y$ for every observation. The sum-of-squares decomposition gives

$$
SS_{tot}=SS_{reg}+SS_{res}.
$$

Consequently, the multiple coefficient of determination is

$$
\boxed{
R^2
=
\frac{SS_{reg}}{SS_{tot}}
=
1-\frac{SS_{res}}{SS_{tot}}.
}
$$

Provided that $SS_{tot}>0$, we have

$$
0\leq R^2\leq 1.
$$

**2. Regression without an intercept**

The intercept-only model is no longer necessarily contained in the regression model. Instead, uncentered $R^2$ uses the zero prediction as its reference:

$$
\boxed{
R_{\mathrm{unc}}^2
=
1-
\frac{SS_{res}}{\mathbf{y}'\mathbf{y}}.
}
$$

Since the zero prediction is always available by setting all coefficients equal to zero, we have

$$
0\leq R_{\mathrm{unc}}^2\leq 1,
$$

provided that $\mathbf{y}'\mathbf{y}>0$.

The usual centered expression, $1-SS_{res}/SS_{tot}$, can be negative for a model without an intercept because its residual sum of squares may exceed that of the intercept-only model.

**3. Interpretation and model comparison**

Both definitions compare the fitted regression model against a reference prediction, but they use different references. Consequently, centered and uncentered $R^2$ should not be compared directly.

An important property of least squares is that adding explanatory variables cannot increase the minimum residual sum of squares. Therefore, when the same observations and reference prediction are used, adding explanatory variables cannot decrease $R^2$.

However, **a higher $R^2$ does not necessarily imply better predictive performance**. It measures improvement in fit to the observations used to estimate the model. To evaluate predictions for new observations, we can instead use methods such as cross-validation.

































## Geometric Interpretation of Multiple Linear Regression

Multiple linear regression can be interpreted geometrically as projecting the response vector $\mathbf{y}$ onto the column space of the design matrix $\mathbf{X}$.

The fitted-value vector $\hat{\mathbf{y}}$ is the orthogonal projection of $\mathbf{y}$ onto $\mathcal{C}(\mathbf{X})$, while the residual vector $\hat{\mathbf{e}}=\mathbf{y}-\hat{\mathbf{y}}$ is orthogonal to this column space.

The following diagram illustrates this interpretation and its connection to the sum-of-squares decomposition.

```{r geo-int, echo=FALSE}
par(mar = c(0,0,0,0))
plot(NULL,
     xlim = c(0,100),
     ylim = c(0,100),
     xaxt = "n",
     yaxt = "n",
     xlab = "",
     ylab = "")
xPad <- 20
yPad <- 50
polygon(x = c(0, xPad, 100, 100 - xPad, 0),
        y = c(0, yPad, yPad,         0, 0),
        col = rgb(0, 0, 0, 0.2),
        border = NA)
segments(x0  = xPad,
         x1  = 100 - xPad - 10,
         y0  = yPad,
         y1  = 100 - 10,
         lwd = 2)
segments(x0  = 100 - xPad - 10,
         x1  = 100 - xPad - 10,
         y0  = 100 - 10,
         y1  = yPad - 20,
         lwd = 2,
         col = 'red')
segments(x0  = xPad,
         x1  = 100 - xPad - 10,
         y0  = yPad,
         y1  = yPad - 20,
         lwd = 2)
segments(x0  = xPad,
         x1  = 100 - xPad - 30,
         y0  = yPad,
         y1  = yPad - 40,
         lwd = 2)
segments(x0  = 100 - xPad - 10,
         x1  = 100 - xPad - 30,
         y0  = 100 - 10,
         y1  = yPad - 40,
         lwd = 2,
         col = 'red')
segments(x0  = 100 - xPad - 10,
         x1  = 100 - xPad - 30,
         y0  = yPad - 20,
         y1  = yPad - 40,
         lwd = 2,
         col = 'red')
segments(x0  = 100 - xPad - 13,
         x1  = 100 - xPad - 13,
         y0  = yPad - 23,
         y1  = yPad - 18,
         lwd = 2,
         col = 'blue')
segments(x0  = 100 - xPad - 13,
         x1  = 100 - xPad - 10,
         y0  = yPad - 18,
         y1  = yPad - 15,
         lwd = 2,
         col = 'blue')
segments(x0  = 100 - xPad - 10,
         x1  = 100 - xPad - 15,
         y0  = yPad - 15,
         y1  = yPad - 13,
         lwd = 2,
         col = 'blue')
segments(x0  = 100 - xPad - 15,
         x1  = 100 - xPad - 15,
         y0  = yPad - 18,
         y1  = yPad - 13,
         lwd = 2,
         col = 'blue')
text(x = 35,
     y = 70,
     labels = expression(y))
text(x = 35,
     y = 25,
     labels = expression(hat(y)[0]))
text(x = 35,
     y = 50,
     labels = expression(hat(y)))
text(x = 35,
     y = 50,
     labels = expression(hat(y)))
text(x = 35,
     y = 50,
     labels = expression(hat(y)))
text(x      = 64,
     y      = 20,
     labels = expression(hat(y) - hat(y)[0]),
     srt    = 40)
text(x      = 73,
     y      = 60,
     labels = expression(y - hat(y)),
     srt    = 90)
text(x      = 55,
     y      = 45,
     labels = expression(y - hat(y)[0]),
     srt    = 70)
text(x = 20,
     y =  5,
     labels = expression(italic("Column Space X")))
text(x = 90,
     y = 90,
     labels = expression(bold(R)^n))
```

The diagram represents the following vectors and subspaces:

- $\mathbf{y}$ is the observed response vector in $\mathbb{R}^n$.
- The gray plane represents the column space $\mathcal{C}(\mathbf{X})$, a subspace of $\mathbb{R}^n$.
- $\hat{\mathbf{y}}$ is the orthogonal projection of $\mathbf{y}$ onto $\mathcal{C}(\mathbf{X})$.
- $\hat{\mathbf{y}}_0=\bar y\mathbf{1}$ is the fitted-value vector from the intercept-only model. Since $\mathbf{X}$ includes an intercept, $\hat{\mathbf{y}}_0$ also belongs to $\mathcal{C}(\mathbf{X})$.

The three vectors

$$
\mathbf{y}-\hat{\mathbf{y}},
\qquad
\hat{\mathbf{y}}-\hat{\mathbf{y}}_0,
\qquad
\mathbf{y}-\hat{\mathbf{y}}_0
$$

form a right triangle.

To see why, notice that both $\hat{\mathbf{y}}$ and $\hat{\mathbf{y}}_0$ belong to $\mathcal{C}(\mathbf{X})$. Therefore,

$$
\hat{\mathbf{y}}-\hat{\mathbf{y}}_0
\in
\mathcal{C}(\mathbf{X}).
$$

From the properties of least squares, we know that the residual vector is orthogonal to every vector in this column space. Consequently,

$$
\underbrace{
(\mathbf{y}-\hat{\mathbf{y}})
}_{\hat{\mathbf{e}}}'
\underbrace{
(\hat{\mathbf{y}}-\hat{\mathbf{y}}_0)
}_{\in\mathcal{C}(\mathbf{X})}
=
0.
$$

Thus, the residual vector and the difference between the two fitted-value vectors are perpendicular. Applying the Pythagorean theorem gives

$$
\left\|
\mathbf{y}-\hat{\mathbf{y}}_0
\right\|^2
=
\left\|
\mathbf{y}-\hat{\mathbf{y}}
\right\|^2
+
\left\|
\hat{\mathbf{y}}-\hat{\mathbf{y}}_0
\right\|^2.
$$

Equivalently,

$$
\begin{aligned}
(\mathbf{y}-\hat{\mathbf{y}}_0)'
(\mathbf{y}-\hat{\mathbf{y}}_0)
&=
(\mathbf{y}-\hat{\mathbf{y}})'
(\mathbf{y}-\hat{\mathbf{y}})
\\
&\quad+
(\hat{\mathbf{y}}-\hat{\mathbf{y}}_0)'
(\hat{\mathbf{y}}-\hat{\mathbf{y}}_0).
\end{aligned}
$$

Since $\hat{\mathbf{y}}_0=\bar y\mathbf{1}$, this is precisely the sum-of-squares decomposition established in the previous section:

$$
\boxed{
SS_{tot}=SS_{res}+SS_{reg}.
}
$$

The geometric interpretation therefore shows that the decomposition is a direct consequence of orthogonal projection, regardless of the number of explanatory variables in the model, provided that an intercept is included.

***































## Centered and Standardized Variables

Centering and standardization extend naturally from simple linear regression to multiple linear regression. They are particularly useful because they provide alternative expressions for the least-squares coefficients in terms of sample covariances and correlations.

Throughout this section, we use the notation

$$
\mathbf{X}_*
=
\begin{bmatrix}
\mathbf{x}_1 & \mathbf{x}_2 & \cdots & \mathbf{x}_p
\end{bmatrix}
\in\mathbb{R}^{n\times p}
$$

for the matrix containing only the explanatory variables. The complete design matrix is

$$
\mathbf{X}
=
\begin{bmatrix}
\mathbf{1} & \mathbf{X}_*
\end{bmatrix}
\in\mathbb{R}^{n\times(p+1)}.
$$

This distinction is useful because the intercept column should not be centered or standardized.

***

### Centered Variables

Let

$$
\bar{\mathbf{x}}
=
\begin{bmatrix}
\bar x_1\\
\bar x_2\\
\vdots\\
\bar x_p
\end{bmatrix}
$$

denote the vector containing the sample mean of each explanatory variable. Since

$$
\bar x_j
=
\frac{1}{n}\mathbf{1}'\mathbf{x}_j,
$$

we can write

\begin{align*}
\bar{\mathbf{x}}
&=
\frac{1}{n}
\begin{bmatrix}
\mathbf{x}_1'\mathbf{1}\\
\mathbf{x}_2'\mathbf{1}\\
\vdots\\
\mathbf{x}_p'\mathbf{1}
\end{bmatrix}
\\
&=
\frac{1}{n}\mathbf{X}_*'\mathbf{1}.
\end{align*}

The centered explanatory-variable matrix is obtained by subtracting the corresponding column mean from every observation:

\begin{align*}
\mathbf{X}_{*,c}
&=
\mathbf{X}_*-\mathbf{1}\bar{\mathbf{x}}'
\\
&=
\mathbf{X}_*
-
\mathbf{1}
\left(
\frac{1}{n}\mathbf{X}_*'\mathbf{1}
\right)'
\\
&=
\mathbf{X}_*
-
\frac{1}{n}
\mathbf{1}\mathbf{1}'\mathbf{X}_*
\\
&=
\left(
\mathbf{I}
-
\frac{1}{n}\mathbf{1}\mathbf{1}'
\right)\mathbf{X}_*.
\end{align*}

This motivates the centering matrix.

```{proposition, prp-mlr-centering-matrix, name="Centering Matrix"}
Define

$$
\mathbf{C}
=
\mathbf{I}
-
\frac{1}{n}\mathbf{1}\mathbf{1}'
=
\mathbf{I}-\mathbf{H}_0,
$$

where

$$
\mathbf{H}_0
=
\frac{1}{n}\mathbf{1}\mathbf{1}'
$$

is the projection matrix associated with the intercept-only model.

Then:

1. $\mathbf{C}$ is symmetric,
2. $\mathbf{C}$ is idempotent,
3. $\mathbf{C}\mathbf{1}=\mathbf{0}$, and
4. for any matrix $\mathbf{A}$ with $n$ rows, $\mathbf{C}\mathbf{A}$ subtracts the column mean from every column of $\mathbf{A}$.

In particular,

$$
\boxed{
\mathbf{X}_{*,c}
=
\mathbf{C}\mathbf{X}_*
}
$$

and

$$
\boxed{
\mathbf{y}_c
=
\mathbf{C}\mathbf{y}
=
\mathbf{y}-\bar y\mathbf{1}.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-centering-matrix}
First,

$$
\mathbf{C}'
=
\left(
\mathbf{I}
-
\frac{1}{n}\mathbf{1}\mathbf{1}'
\right)'
=
\mathbf{I}
-
\frac{1}{n}\mathbf{1}\mathbf{1}'
=
\mathbf{C},
$$

so $\mathbf{C}$ is symmetric.

Also,

\begin{align*}
\mathbf{C}^2
&=
\left(
\mathbf{I}
-
\frac{1}{n}\mathbf{1}\mathbf{1}'
\right)^2
\\
&=
\mathbf{I}
-
\frac{2}{n}\mathbf{1}\mathbf{1}'
+
\frac{1}{n^2}
\mathbf{1}\mathbf{1}'\mathbf{1}\mathbf{1}'
\\
&=
\mathbf{I}
-
\frac{2}{n}\mathbf{1}\mathbf{1}'
+
\frac{1}{n}
\mathbf{1}\mathbf{1}'
\\
&=
\mathbf{I}
-
\frac{1}{n}\mathbf{1}\mathbf{1}'
\\
&=
\mathbf{C},
\end{align*}

where we used $\mathbf{1}'\mathbf{1}=n$. Thus, $\mathbf{C}$ is idempotent.

Similarly,

\begin{align*}
\mathbf{C}\mathbf{1}
&=
\left(
\mathbf{I}
-
\frac{1}{n}\mathbf{1}\mathbf{1}'
\right)\mathbf{1}
\\
&=
\mathbf{1}
-
\frac{1}{n}\mathbf{1}
(\mathbf{1}'\mathbf{1})
\\
&=
\mathbf{1}-\mathbf{1}
\\
&=
\mathbf{0}.
\end{align*}

Finally, if $\mathbf{A}$ has $n$ rows, then

$$
\frac{1}{n}\mathbf{1}'\mathbf{A}
$$

contains the means of its columns. Therefore,

$$
\mathbf{C}\mathbf{A}
=
\mathbf{A}
-
\mathbf{1}
\left(
\frac{1}{n}\mathbf{1}'\mathbf{A}
\right)
$$

subtracts the corresponding mean from each column.
```

</details>

***

### Sample Covariance

Using the centered explanatory variables, define the sample covariance matrix

$$
\mathbf{S}_{XX}
\in
\mathbb{R}^{p\times p}
$$

by

$$
\mathbf{S}_{XX}
=
\frac{1}{n-1}
\mathbf{X}_{*,c}'\mathbf{X}_{*,c}.
$$

Its $(j,k)$ entry is the sample covariance between explanatory variables $\mathbf{x}_j$ and $\mathbf{x}_k$.

Similarly, define the covariance vector between the explanatory variables and the response as

$$
\mathbf{S}_{Xy}
=
\frac{1}{n-1}
\mathbf{X}_{*,c}'\mathbf{y}_c
\in
\mathbb{R}^{p}.
$$

The centering matrix gives several useful equivalent expressions.

```{proposition, prp-mlr-covariance-identities, name="Covariance Identities"}
The sample covariance matrix satisfies

$$
\boxed{
\mathbf{S}_{XX}
=
\frac{1}{n-1}
\mathbf{X}_*'\mathbf{C}\mathbf{X}_*
=
\frac{1}{n-1}
\mathbf{X}_{*,c}'\mathbf{X}_*
=
\frac{1}{n-1}
\mathbf{X}_*'\mathbf{X}_{*,c}.
}
$$

Similarly,

$$
\boxed{
\mathbf{S}_{Xy}
=
\frac{1}{n-1}
\mathbf{X}_*'\mathbf{C}\mathbf{y}
=
\frac{1}{n-1}
\mathbf{X}_{*,c}'\mathbf{y}
=
\frac{1}{n-1}
\mathbf{X}_*'\mathbf{y}_c.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-covariance-identities}
Since

$$
\mathbf{X}_{*,c}
=
\mathbf{C}\mathbf{X}_*,
$$

we have

\begin{align*}
\mathbf{S}_{XX}
&=
\frac{1}{n-1}
\mathbf{X}_{*,c}'\mathbf{X}_{*,c}
\\
&=
\frac{1}{n-1}
\mathbf{X}_*'\mathbf{C}'\mathbf{C}\mathbf{X}_*
\\
&=
\frac{1}{n-1}
\mathbf{X}_*'\mathbf{C}\mathbf{X}_*,
\end{align*}

because $\mathbf{C}$ is symmetric and idempotent.

Also,

$$
\mathbf{X}_*'\mathbf{C}\mathbf{X}_*
=
\mathbf{X}_{*,c}'\mathbf{X}_*
=
\mathbf{X}_*'\mathbf{X}_{*,c}.
$$

The covariance vector follows in the same way:

\begin{align*}
\mathbf{S}_{Xy}
&=
\frac{1}{n-1}
\mathbf{X}_{*,c}'\mathbf{y}_c
\\
&=
\frac{1}{n-1}
\mathbf{X}_*'\mathbf{C}'\mathbf{C}\mathbf{y}
\\
&=
\frac{1}{n-1}
\mathbf{X}_*'\mathbf{C}\mathbf{y}.
\end{align*}

Therefore,

$$
\mathbf{S}_{Xy}
=
\frac{1}{n-1}\mathbf{X}_{*,c}'\mathbf{y}
=
\frac{1}{n-1}\mathbf{X}_*'\mathbf{y}_c.
$$
```

</details>

Thus, for the covariance calculation, centering either side is sufficient because multiplication by the centering matrix removes the corresponding mean automatically.

***

### Least-Squares Coefficients in Covariance Form

We now use the covariance quantities to obtain an alternative expression for the multiple regression coefficients.

Write the coefficient vector as

$$
\hat{\boldsymbol{\beta}}
=
\begin{bmatrix}
\hat\beta_0\\
\hat{\boldsymbol{\beta}}_{-0}
\end{bmatrix},
$$

where

$$
\hat{\boldsymbol{\beta}}_{-0}
=
\begin{bmatrix}
\hat\beta_1\\
\vdots\\
\hat\beta_p
\end{bmatrix}
$$

contains the coefficients associated with the explanatory variables.

Since the complete design matrix is

$$
\mathbf{X}
=
\begin{bmatrix}
\mathbf{1} & \mathbf{X}_*
\end{bmatrix},
$$

the usual least-squares solution is

$$
\hat{\boldsymbol{\beta}}
=
(\mathbf{X}'\mathbf{X})^{-1}
\mathbf{X}'\mathbf{y}.
$$

The covariance representation follows by explicitly evaluating this expression.

```{proposition, prp-mlr-coefficients-covariance, name="Regression Coefficients in Covariance Form"}
Suppose that the design matrix

$$
\mathbf{X}
=
\begin{bmatrix}
\mathbf{1} & \mathbf{X}_*
\end{bmatrix}
$$

has full column rank.

Then the estimated coefficients associated with the explanatory variables satisfy

$$
\boxed{
\hat{\boldsymbol{\beta}}_{-0}
=
\mathbf{S}_{XX}^{-1}
\mathbf{S}_{Xy}.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-coefficients-covariance}
We begin from

$$
\hat{\boldsymbol{\beta}}
=
(\mathbf{X}'\mathbf{X})^{-1}
\mathbf{X}'\mathbf{y}.
$$

Since

$$
\mathbf{X}
=
\begin{bmatrix}
\mathbf{1} & \mathbf{X}_*
\end{bmatrix},
$$

we have

\begin{align*}
\mathbf{X}'\mathbf{X}
&=
\begin{bmatrix}
\mathbf{1}'\\
\mathbf{X}_*'
\end{bmatrix}
\begin{bmatrix}
\mathbf{1} & \mathbf{X}_*
\end{bmatrix}
\\
&=
\begin{bmatrix}
\mathbf{1}'\mathbf{1}
&
\mathbf{1}'\mathbf{X}_*
\\
\mathbf{X}_*'\mathbf{1}
&
\mathbf{X}_*'\mathbf{X}_*
\end{bmatrix}.
\end{align*}

Because

$$
\mathbf{1}'\mathbf{1}=n
$$

and

$$
\bar{\mathbf{x}}
=
\frac{1}{n}\mathbf{X}_*'\mathbf{1},
$$

we obtain

$$
\mathbf{X}'\mathbf{X}
=
\begin{bmatrix}
n
&
n\bar{\mathbf{x}}'
\\
n\bar{\mathbf{x}}
&
\mathbf{X}_*'\mathbf{X}_*
\end{bmatrix}.
$$

To invert this matrix, use the block-matrix inverse with upper-left block $n$. The corresponding Schur complement is

\begin{align*}
\mathbf{X}_*'\mathbf{X}_*
-
(n\bar{\mathbf{x}})
n^{-1}
(n\bar{\mathbf{x}}')
&=
\mathbf{X}_*'\mathbf{X}_*
-
n\bar{\mathbf{x}}\bar{\mathbf{x}}'
\\
&=
\mathbf{X}_*'\mathbf{X}_*
-
\frac{1}{n}
\mathbf{X}_*'\mathbf{1}\mathbf{1}'\mathbf{X}_*
\\
&=
\mathbf{X}_*'
\left(
\mathbf{I}
-
\frac{1}{n}\mathbf{1}\mathbf{1}'
\right)
\mathbf{X}_*
\\
&=
\mathbf{X}_*'\mathbf{C}\mathbf{X}_*
\\
&=
(n-1)\mathbf{S}_{XX}.
\end{align*}

Therefore,

\begin{align*}
(\mathbf{X}'\mathbf{X})^{-1}
&=
\begin{bmatrix}
\displaystyle
\frac{1}{n}
+
\bar{\mathbf{x}}'
\bigl((n-1)\mathbf{S}_{XX}\bigr)^{-1}
\bar{\mathbf{x}}
&
\displaystyle
-\bar{\mathbf{x}}'
\bigl((n-1)\mathbf{S}_{XX}\bigr)^{-1}
\\[8pt]
\displaystyle
-\bigl((n-1)\mathbf{S}_{XX}\bigr)^{-1}
\bar{\mathbf{x}}
&
\displaystyle
\bigl((n-1)\mathbf{S}_{XX}\bigr)^{-1}
\end{bmatrix}
\\
&=
\begin{bmatrix}
\displaystyle
\frac{1}{n}
+
\frac{1}{n-1}
\bar{\mathbf{x}}'
\mathbf{S}_{XX}^{-1}
\bar{\mathbf{x}}
&
\displaystyle
-\frac{1}{n-1}
\bar{\mathbf{x}}'
\mathbf{S}_{XX}^{-1}
\\[8pt]
\displaystyle
-\frac{1}{n-1}
\mathbf{S}_{XX}^{-1}
\bar{\mathbf{x}}
&
\displaystyle
\frac{1}{n-1}
\mathbf{S}_{XX}^{-1}
\end{bmatrix}.
\end{align*}

Also,

\begin{align*}
\mathbf{X}'\mathbf{y}
&=
\begin{bmatrix}
\mathbf{1}'\mathbf{y}\\
\mathbf{X}_*'\mathbf{y}
\end{bmatrix}
\\
&=
\begin{bmatrix}
n\bar y\\
\mathbf{X}_*'\mathbf{y}
\end{bmatrix}.
\end{align*}

Multiplying the second row-block gives

\begin{align*}
\hat{\boldsymbol{\beta}}_{-0}
&=
-\frac{n\bar y}{n-1}
\mathbf{S}_{XX}^{-1}\bar{\mathbf{x}}
+
\frac{1}{n-1}
\mathbf{S}_{XX}^{-1}\mathbf{X}_*'\mathbf{y}
\\
&=
\frac{1}{n-1}
\mathbf{S}_{XX}^{-1}
\left(
\mathbf{X}_*'\mathbf{y}
-
n\bar{\mathbf{x}}\bar y
\right).
\end{align*}

Now,

\begin{align*}
\mathbf{X}_*'\mathbf{y}
-
n\bar{\mathbf{x}}\bar y
&=
\mathbf{X}_*'\mathbf{y}
-
\frac{1}{n}
\mathbf{X}_*'\mathbf{1}\mathbf{1}'\mathbf{y}
\\
&=
\mathbf{X}_*'
\left(
\mathbf{I}
-
\frac{1}{n}\mathbf{1}\mathbf{1}'
\right)
\mathbf{y}
\\
&=
\mathbf{X}_*'\mathbf{C}\mathbf{y}
\\
&=
(n-1)\mathbf{S}_{Xy}.
\end{align*}

Therefore,

\begin{align*}
\hat{\boldsymbol{\beta}}_{-0}
&=
\frac{1}{n-1}
\mathbf{S}_{XX}^{-1}
(n-1)\mathbf{S}_{Xy}
\\
&=
\mathbf{S}_{XX}^{-1}\mathbf{S}_{Xy}.
\end{align*}

Hence,

$$
\boxed{
\hat{\boldsymbol{\beta}}_{-0}
=
\mathbf{S}_{XX}^{-1}\mathbf{S}_{Xy}.
}
$$
```

</details>

This expression separates two components of the multiple-regression problem:

- $\mathbf{S}_{XX}^{-1}$ incorporates the relationships among the explanatory variables.
- $\mathbf{S}_{Xy}$ incorporates the relationships between the explanatory variables and the response.

Thus, unlike simple linear regression, each coefficient depends not only on the relationship between one explanatory variable and the response, but also on how the explanatory variables relate to one another.

***

### The Intercept and the Effect of Centering

The corresponding intercept estimate also has a convenient covariance-form expression.

```{proposition, prp-mlr-intercept-covariance, name="Intercept Estimate"}
Under the assumptions of the previous proposition,

$$
\boxed{
\hat\beta_0
=
\bar y
-
\bar{\mathbf{x}}'
\hat{\boldsymbol{\beta}}_{-0}.
}
$$

Equivalently,

$$
\boxed{
\hat\beta_0
=
\bar y
-
\bar{\mathbf{x}}'
\mathbf{S}_{XX}^{-1}
\mathbf{S}_{Xy}.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-intercept-covariance}
Using the first row-block of the block-matrix expression,

\begin{align*}
\hat\beta_0
&=
\bar y
+
\frac{n\bar y}{n-1}
\bar{\mathbf{x}}'
\mathbf{S}_{XX}^{-1}
\bar{\mathbf{x}}
-
\frac{1}{n-1}
\bar{\mathbf{x}}'
\mathbf{S}_{XX}^{-1}
\mathbf{X}_*'\mathbf{y}
\\
&=
\bar y
-
\frac{1}{n-1}
\bar{\mathbf{x}}'
\mathbf{S}_{XX}^{-1}
\left(
\mathbf{X}_*'\mathbf{y}
-
n\bar{\mathbf{x}}\bar y
\right).
\end{align*}

From the previous proof,

$$
\mathbf{X}_*'\mathbf{y}
-
n\bar{\mathbf{x}}\bar y
=
(n-1)\mathbf{S}_{Xy}.
$$

Therefore,

\begin{align*}
\hat\beta_0
&=
\bar y
-
\bar{\mathbf{x}}'
\mathbf{S}_{XX}^{-1}\mathbf{S}_{Xy}
\\
&=
\bar y
-
\bar{\mathbf{x}}'
\hat{\boldsymbol{\beta}}_{-0}.
\end{align*}
```

</details>

These results also clarify what happens when variables are centered.

```{proposition, prp-mlr-centering-coefficients, name="Effect of Centering on the Regression Coefficients"}
Suppose the regression model includes an intercept.

1. Centering the explanatory variables does not change their estimated coefficients.

2. If only the explanatory variables are centered, the estimated intercept becomes

$$
\boxed{
\hat\beta_{0,c}=\bar y.
}
$$

3. If both the explanatory variables and the response are centered, the estimated intercept becomes

$$
\boxed{
\hat\beta_{0,c}=0.
}
$$

In the third case, fitting the centered variables without an intercept produces the same slope estimates and fitted centered values as fitting them with an intercept.
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-centering-coefficients}
Centering does not change the covariance matrix or covariance vector:

$$
\mathbf{S}_{X_cX_c}
=
\mathbf{S}_{XX},
\qquad
\mathbf{S}_{X_cy}
=
\mathbf{S}_{Xy}.
$$

Therefore,

$$
\hat{\boldsymbol{\beta}}_{-0,c}
=
\mathbf{S}_{X_cX_c}^{-1}
\mathbf{S}_{X_cy}
=
\mathbf{S}_{XX}^{-1}
\mathbf{S}_{Xy}
=
\hat{\boldsymbol{\beta}}_{-0}.
$$

If the explanatory variables are centered, their mean vector is $\mathbf{0}$. Hence,

$$
\hat\beta_{0,c}
=
\bar y
-
\mathbf{0}'
\hat{\boldsymbol{\beta}}_{-0}
=
\bar y.
$$

If the response is also centered, then $\bar y_c=0$, and therefore

$$
\hat\beta_{0,c}
=
0.
$$

Thus, when both the response and explanatory variables are centered, the intercept estimated by a model that includes one is exactly zero. Removing the intercept in this centered problem therefore leaves the slope estimates unchanged.
```

</details>

The last statement applies specifically to the regression of the **centered response on the centered explanatory variables**. It does not imply that an intercept can generally be omitted from a regression on the original, uncentered variables.

***

### Standardized Variables

Centering removes location. Standardization additionally rescales each variable by its sample standard deviation.

Assume that every explanatory variable has positive sample variance. Define

$$
\mathbf{D}_X
=
\begin{bmatrix}
S_{x_1x_1}^{-1/2} & 0 & \cdots & 0\\
0 & S_{x_2x_2}^{-1/2} & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots\\
0 & 0 & \cdots & S_{x_px_p}^{-1/2}
\end{bmatrix}.
$$

The standardized explanatory-variable matrix is

$$
\boxed{
\mathbf{X}_s
=
\mathbf{X}_{*,c}\mathbf{D}_X
=
\mathbf{C}\mathbf{X}_*\mathbf{D}_X.
}
$$

Thus, the $j$th column of $\mathbf{X}_s$ is

$$
\frac{\mathbf{x}_j-\bar x_j\mathbf{1}}
{\sqrt{S_{x_jx_j}}}.
$$

The matrix $\mathbf{C}$ acts on the rows and can center any matrix with $n$ rows. In contrast, $\mathbf{D}_X$ acts on the columns and is specific to the standard deviations of the explanatory variables in $\mathbf{X}_*$.

Similarly, if $S_{yy}>0$, the standardized response is

$$
\mathbf{y}_s
=
\frac{\mathbf{y}-\bar y\mathbf{1}}
{\sqrt{S_{yy}}}.
$$

Because standardized variables are centered, their sample means are zero.

***

### Sample Correlation Matrix

The sample correlation matrix of the explanatory variables is defined by

$$
\mathbf{r}_{XX}
=
\frac{1}{n-1}
\mathbf{X}_s'\mathbf{X}_s.
$$

Its $(j,k)$ entry is

$$
[\mathbf{r}_{XX}]_{jk}
=
\frac{S_{x_jx_k}}
{\sqrt{S_{x_jx_j}S_{x_kx_k}}},
$$

the sample correlation between explanatory variables $\mathbf{x}_j$ and $\mathbf{x}_k$.

Similarly, define the sample correlation vector between the explanatory variables and the response as

$$
\mathbf{r}_{Xy}
=
\frac{1}{n-1}
\mathbf{X}_s'\mathbf{y}_s.
$$

Equivalently,

$$
\mathbf{r}_{XX}
=
\mathbf{D}_X
\mathbf{S}_{XX}
\mathbf{D}_X,
$$

and

$$
\mathbf{r}_{Xy}
=
\frac{1}{\sqrt{S_{yy}}}
\mathbf{D}_X
\mathbf{S}_{Xy}.
$$

***

### Regression with Standardized Variables

Since both $\mathbf{X}_s$ and $\mathbf{y}_s$ are centered, fitting the standardized regression with an intercept produces an estimated intercept of zero. Consequently, the same slope estimates can be obtained by fitting

$$
\mathbf{y}_s
=
\mathbf{X}_s\boldsymbol{\beta}_s
+
\mathbf{e}_s
$$

without explicitly including an intercept.

```{proposition, prp-mlr-standardized-coefficients, name="Regression Coefficients for Standardized Variables"}
Suppose that $\mathbf{r}_{XX}$ is invertible. The least-squares coefficients from the regression of the standardized response on the standardized explanatory variables satisfy

$$
\boxed{
\hat{\boldsymbol{\beta}}_s
=
\mathbf{r}_{XX}^{-1}
\mathbf{r}_{Xy}.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-standardized-coefficients}
The least-squares solution is

$$
\hat{\boldsymbol{\beta}}_s
=
(\mathbf{X}_s'\mathbf{X}_s)^{-1}
\mathbf{X}_s'\mathbf{y}_s.
$$

From the definitions of the sample correlation matrix and correlation vector,

$$
\mathbf{X}_s'\mathbf{X}_s
=
(n-1)\mathbf{r}_{XX}
$$

and

$$
\mathbf{X}_s'\mathbf{y}_s
=
(n-1)\mathbf{r}_{Xy}.
$$

Therefore,

\begin{align*}
\hat{\boldsymbol{\beta}}_s
&=
\left(
(n-1)\mathbf{r}_{XX}
\right)^{-1}
(n-1)\mathbf{r}_{Xy}
\\
&=
\mathbf{r}_{XX}^{-1}
\mathbf{r}_{Xy}.
\end{align*}
```

</details>

This result is the standardized analogue of

$$
\hat{\boldsymbol{\beta}}_{-0}
=
\mathbf{S}_{XX}^{-1}\mathbf{S}_{Xy}.
$$

For the original variables, the slope coefficients depend on the covariance structure. For standardized variables, they depend on the corresponding correlation structure.

Because standardized variables are unitless and each has sample standard deviation one, the magnitudes of their coefficients are more directly comparable than coefficients measured on different physical scales. However, coefficient magnitude should not by itself be interpreted as a measure of variable importance. The coefficients still depend on the relationships among all explanatory variables through $\mathbf{r}_{XX}^{-1}$.

***

### Example: GDP Data

We can verify these results using the GDP data introduced earlier.

The notation used in the R code differs slightly from the mathematical notation above:

- `X` corresponds to the explanatory-variable matrix $\mathbf{X}_*$.
- `Z` corresponds to the complete design matrix $\mathbf{X}$.
- `Xc` and `yc` correspond to $\mathbf{X}_{*,c}$ and $\mathbf{y}_c$.

The first code block constructs the centering matrix and compares coefficient estimates under several centering choices.

```{r gdp-centering}
# Read Data
dat <- read.csv("Gdp data.csv")

# Design Matrix Independent Variables
X <- as.matrix(dat[, -1])
# Dependent Variable
y <- dat$gdp
# Number of Observations
n <- nrow(X)
# Design Matrix with the column of Ones
Z <- cbind(rep(1, n), X)

# Centering
# Vector of Ones
v1 <- rep(1, n)
# Centering Matrix
C  <- diag(n) - (1/n) * v1 %*% t(v1)
# Independent Variables Centered
Xc <- C %*% X
# Dependent Variable Centered
yc <- C %*% y
# Design Matrix with Independent Variables Centered
Zc <- cbind(rep(1, n), Xc)

# Checks that the Variables are actually centered
print(round(colMeans(Xc), 8))
print(round(mean(yc), 8))

# Shows the Mean of y
print(round(mean(y), 8))

# Compute the Estimates of the Coefficients with Original Variables
b   <- solve(t(Z) %*% Z, t(Z) %*% y)
# Compute the Estimates of the Coefficients with Centered Independent Variables Only
bs1 <- solve(t(Zc) %*% Zc, t(Zc) %*% y)
# Compute the Estimates of the Coefficients with All Variables Centered
bs2 <- solve(t(Zc) %*% Zc, t(Zc) %*% yc)
# Compute the Estimates of the Coefficients with All variables Centered and no column of ones
bs3 <- solve(t(Xc) %*% Xc, t(Xc) %*% yc)

# Shows the Estimated Coefficients Side-by-Side
tab <- cbind(b, bs1, bs2, c(0, bs3))
colnames(tab) <- c("Original", "Center Exp", "Both Centered", "No Intercept")
print(round(tab, 8))
```

The output illustrates the theoretical results:

- Centering the explanatory variables does not change their slope estimates.
- When only the explanatory variables are centered, the intercept becomes $\bar y$.
- When both the response and explanatory variables are centered, the intercept becomes zero.
- Fitting the centered response on the centered explanatory variables without an intercept produces the same slope estimates.

We can next verify the covariance formulas directly. The following code computes $\mathbf{S}_{XX}$ and $\mathbf{S}_{Xy}$ from the centered variables and compares them with the corresponding built-in R calculations.

```{r gdp-sample-covariance}
# Sample Covariance of X
SXX <- t(Xc) %*% Xc / (n-1)
# Sample Covariance between X and y
SXy <- t(Xc) %*% yc / (n-1)

# Shows the comparison in covariance matrices
print(SXX)
print(cov(X))

# Shows the comparison in covariance vectors
print(SXy)
print(cov(X, y))
```

The two calculations should agree, up to numerical precision.

We can also compute the regression coefficients directly from the covariance identities,

$$
\hat{\boldsymbol{\beta}}_{-0}
=
\mathbf{S}_{XX}^{-1}\mathbf{S}_{Xy}
$$

and

$$
\hat\beta_0
=
\bar y
-
\bar{\mathbf{x}}'
\hat{\boldsymbol{\beta}}_{-0}.
$$

```{r gdp-sample-covariance-estimation}
# Computes the estimates of the coefficients using the covariance matrices
b1 <- solve(SXX, SXy)
b0 <- mean(y) - t(colMeans(X)) %*% b1

# Shows the Estimates Side by Side
tab <- cbind(b, c(b0, b1))
colnames(tab) <- c("Regular", "Split")
print(round(cbind(b, c(b0, b1)), 8))
```

The resulting estimates agree with those obtained from the ordinary least-squares calculation.

We now standardize the explanatory variables and response. In the code, `DX` corresponds to $\mathbf{D}_X$, `Xs` to $\mathbf{X}_s$, and `ys` to $\mathbf{y}_s$.

```{r gdp-sample-correlation}
# Standardizing matrix of X
DX <- diag(1/sqrt(diag(SXX)))
# Standardize X
Xs <- Xc %*% DX
# Shows that Xs is indeed standardize
print(round(colMeans(Xs), 8))
print(apply(X = Xs, MARGIN = 2, FUN = sd))

# Standardizes y
ys <- yc / sd(y)
# Shows that ys is standardized
print(round(mean(ys), 8))
print(round(sd(ys), 8))

# Sample Correlation of X
rXX <- t(Xs) %*% Xs / (n-1)
# Sample Correlation between X and y
rXy <- t(Xs) %*% ys / (n-1)

# Shows the comparison in correlation matrices
print(rXX)
print(cor(X))

# Shows the comparison in correlation vectors
print(rXy)
print(cor(X, y))
```

The standardized explanatory variables have sample mean zero and sample standard deviation one. The quantities computed from their cross-products agree with the sample correlations produced by `cor`.

Finally, we verify

$$
\hat{\boldsymbol{\beta}}_s
=
\mathbf{r}_{XX}^{-1}
\mathbf{r}_{Xy}.
$$

```{r gdp-sample-correlation-estimation}
# Computes the estimates of the coefficients using the covariance matrices
bs1 <- solve(t(Xs) %*% Xs, t(Xs) %*% ys)
bs2 <- solve(rXX, rXy)

# Shows the Estimates Side by Side
print(round(cbind(bs1, bs2), 8))
```

Both calculations produce the same standardized slope estimates.

We can also compare the coefficients obtained on the original and standardized scales.

```{r gdp-sample-standardization-comparisson}
# Computes the estimates of the coefficients using the covariance matrices
bs1 <- lm(y ~ X)$coefficients
bs2 <- lm(ys ~ Xs)$coefficients

# Shows the Estimates Side by Side
print(round(cbind(bs1, bs2), 8))
```

Since `ys` and every column of `Xs` are centered, the intercept estimated by `lm(ys ~ Xs)` is zero up to numerical precision. We could therefore obtain the same standardized slope estimates using `lm(ys ~ Xs - 1)`.

The remaining coefficients generally change in magnitude because each variable is now measured in standard-deviation units rather than its original units. This removes differences caused purely by measurement scale, while the relationships among the explanatory variables continue to affect the coefficient estimates.

***



























## Variable Cross-Effects

In the previous section, we expressed the regression coefficients for standardized variables in terms of the sample correlation matrix. We now use this representation to study how relationships among the explanatory variables affect their estimated coefficients.

Because all variables in this section are standardized, they are centered. Therefore, the intercept estimate is zero, and we can work with the standardized explanatory variables without explicitly including a column of ones.

Let the standardized explanatory-variable matrix be partitioned into two groups,

$$
\mathbf{X}_s
=
\begin{bmatrix}
\mathbf{X}_1^s & \mathbf{X}_2^s
\end{bmatrix},
$$

and partition the corresponding coefficient vector as

$$
\hat{\boldsymbol{\beta}}_s
=
\begin{bmatrix}
\hat{\boldsymbol{\beta}}_1^s\\
\hat{\boldsymbol{\beta}}_2^s
\end{bmatrix}.
$$

The objective is to understand how the variables in $\mathbf{X}_2^s$ affect the coefficients associated with $\mathbf{X}_1^s$, and vice versa.

***

### Block Correlation Representation

Define the correlation blocks

$$
\mathbf{R}_{11}
=
\frac{1}{n-1}
(\mathbf{X}_1^s)'\mathbf{X}_1^s,
$$

$$
\mathbf{R}_{12}
=
\frac{1}{n-1}
(\mathbf{X}_1^s)'\mathbf{X}_2^s,
$$

$$
\mathbf{R}_{21}
=
\frac{1}{n-1}
(\mathbf{X}_2^s)'\mathbf{X}_1^s,
$$

and

$$
\mathbf{R}_{22}
=
\frac{1}{n-1}
(\mathbf{X}_2^s)'\mathbf{X}_2^s.
$$

Since these are sample correlations,

$$
\mathbf{R}_{21}
=
\mathbf{R}_{12}'.
$$

Similarly, define the correlation vectors with the standardized response $\mathbf{y}_s$ as

$$
\mathbf{r}_{1y}
=
\frac{1}{n-1}
(\mathbf{X}_1^s)'\mathbf{y}_s
$$

and

$$
\mathbf{r}_{2y}
=
\frac{1}{n-1}
(\mathbf{X}_2^s)'\mathbf{y}_s.
$$

```{proposition, prp-mlr-block-correlation-estimates, name="Block Correlation Representation of the Coefficient Estimates"}
The least-squares estimates based on the standardized variables satisfy

$$
\boxed{
\begin{bmatrix}
\hat{\boldsymbol{\beta}}_1^s\\
\hat{\boldsymbol{\beta}}_2^s
\end{bmatrix}
=
\begin{bmatrix}
\mathbf{R}_{11} & \mathbf{R}_{12}\\
\mathbf{R}_{21} & \mathbf{R}_{22}
\end{bmatrix}^{-1}
\begin{bmatrix}
\mathbf{r}_{1y}\\
\mathbf{r}_{2y}
\end{bmatrix}.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-block-correlation-estimates}
The least-squares solution for the standardized variables is

$$
\hat{\boldsymbol{\beta}}_s
=
(\mathbf{X}_s'\mathbf{X}_s)^{-1}
\mathbf{X}_s'\mathbf{y}_s.
$$

Using the partition

$$
\mathbf{X}_s
=
\begin{bmatrix}
\mathbf{X}_1^s & \mathbf{X}_2^s
\end{bmatrix},
$$

we have

\begin{align*}
\mathbf{X}_s'\mathbf{X}_s
&=
\begin{bmatrix}
(\mathbf{X}_1^s)'\\
(\mathbf{X}_2^s)'
\end{bmatrix}
\begin{bmatrix}
\mathbf{X}_1^s & \mathbf{X}_2^s
\end{bmatrix}
\\
&=
\begin{bmatrix}
(\mathbf{X}_1^s)'\mathbf{X}_1^s
&
(\mathbf{X}_1^s)'\mathbf{X}_2^s
\\
(\mathbf{X}_2^s)'\mathbf{X}_1^s
&
(\mathbf{X}_2^s)'\mathbf{X}_2^s
\end{bmatrix}
\\
&=
(n-1)
\begin{bmatrix}
\mathbf{R}_{11} & \mathbf{R}_{12}\\
\mathbf{R}_{21} & \mathbf{R}_{22}
\end{bmatrix}.
\end{align*}

Similarly,

\begin{align*}
\mathbf{X}_s'\mathbf{y}_s
&=
\begin{bmatrix}
(\mathbf{X}_1^s)'\mathbf{y}_s\\
(\mathbf{X}_2^s)'\mathbf{y}_s
\end{bmatrix}
\\
&=
(n-1)
\begin{bmatrix}
\mathbf{r}_{1y}\\
\mathbf{r}_{2y}
\end{bmatrix}.
\end{align*}

Therefore,

\begin{align*}
\hat{\boldsymbol{\beta}}_s
&=
\left(
(n-1)
\begin{bmatrix}
\mathbf{R}_{11} & \mathbf{R}_{12}\\
\mathbf{R}_{21} & \mathbf{R}_{22}
\end{bmatrix}
\right)^{-1}
(n-1)
\begin{bmatrix}
\mathbf{r}_{1y}\\
\mathbf{r}_{2y}
\end{bmatrix}
\\
&=
\begin{bmatrix}
\mathbf{R}_{11} & \mathbf{R}_{12}\\
\mathbf{R}_{21} & \mathbf{R}_{22}
\end{bmatrix}^{-1}
\begin{bmatrix}
\mathbf{r}_{1y}\\
\mathbf{r}_{2y}
\end{bmatrix}.
\end{align*}
```

</details>

The coefficient estimates therefore depend jointly on the correlations between the response and each group and on the correlations between the two groups of explanatory variables.

***

### Coefficients for One Group After Accounting for the Other

To study the coefficient estimates for group 1, define

$$
\boxed{
\mathbf{R}_{1\mid2}
=
\mathbf{R}_{11}
-
\mathbf{R}_{12}
\mathbf{R}_{22}^{-1}
\mathbf{R}_{21}.
}
$$

This matrix is the **Schur complement** of $\mathbf{R}_{22}$ in the full correlation matrix.

We assume that $\mathbf{R}_{22}$ is invertible and that $\mathbf{R}_{1\mid2}$ is invertible. These conditions hold, in particular, when the complete standardized design matrix has full column rank.

The matrix $\mathbf{R}_{1\mid2}$ describes the remaining variation and relationships within group 1 after its linear relationships with group 2 have been removed.

```{proposition, prp-mlr-group-cross-effects, name="Coefficient Estimates After Accounting for Another Group"}
Under the invertibility conditions above,

$$
\boxed{
\hat{\boldsymbol{\beta}}_1^s
=
\mathbf{R}_{1\mid2}^{-1}
\left(
\mathbf{r}_{1y}
-
\mathbf{R}_{12}
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2y}
\right).
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-group-cross-effects}
Consider the block correlation matrix

$$
\mathbf{R}
=
\begin{bmatrix}
\mathbf{R}_{11} & \mathbf{R}_{12}\\
\mathbf{R}_{21} & \mathbf{R}_{22}
\end{bmatrix}.
$$

Using the block-matrix inverse with $\mathbf{R}_{22}$ as the invertible block and

$$
\mathbf{R}_{1\mid2}
=
\mathbf{R}_{11}
-
\mathbf{R}_{12}
\mathbf{R}_{22}^{-1}
\mathbf{R}_{21},
$$

we obtain

$$
\mathbf{R}^{-1}
=
\begin{bmatrix}
\mathbf{R}_{1\mid2}^{-1}
&
-\mathbf{R}_{1\mid2}^{-1}
\mathbf{R}_{12}
\mathbf{R}_{22}^{-1}
\\[6pt]
-\mathbf{R}_{22}^{-1}
\mathbf{R}_{21}
\mathbf{R}_{1\mid2}^{-1}
&
\mathbf{R}_{22}^{-1}
+
\mathbf{R}_{22}^{-1}
\mathbf{R}_{21}
\mathbf{R}_{1\mid2}^{-1}
\mathbf{R}_{12}
\mathbf{R}_{22}^{-1}
\end{bmatrix}.
$$

Therefore,

\begin{align*}
\begin{bmatrix}
\hat{\boldsymbol{\beta}}_1^s\\
\hat{\boldsymbol{\beta}}_2^s
\end{bmatrix}
&=
\mathbf{R}^{-1}
\begin{bmatrix}
\mathbf{r}_{1y}\\
\mathbf{r}_{2y}
\end{bmatrix}.
\end{align*}

The first row-block gives

\begin{align*}
\hat{\boldsymbol{\beta}}_1^s
&=
\mathbf{R}_{1\mid2}^{-1}\mathbf{r}_{1y}
-
\mathbf{R}_{1\mid2}^{-1}
\mathbf{R}_{12}
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2y}
\\
&=
\mathbf{R}_{1\mid2}^{-1}
\left(
\mathbf{r}_{1y}
-
\mathbf{R}_{12}
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2y}
\right).
\end{align*}

Thus,

$$
\boxed{
\hat{\boldsymbol{\beta}}_1^s
=
\mathbf{R}_{1\mid2}^{-1}
\left(
\mathbf{r}_{1y}
-
\mathbf{R}_{12}
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2y}
\right).
}
$$
```

</details>

Because the two groups play symmetric roles, an analogous expression can be obtained for $\hat{\boldsymbol{\beta}}_2^s$ by reversing the groups.

***

### Full Model Versus Group-Only Models

Suppose instead that we fit two separate regressions: one using only the variables in group 1 and another using only those in group 2.

The resulting coefficient estimates are

$$
\tilde{\boldsymbol{\beta}}_1^s
=
\mathbf{R}_{11}^{-1}\mathbf{r}_{1y}
$$

and

$$
\tilde{\boldsymbol{\beta}}_2^s
=
\mathbf{R}_{22}^{-1}\mathbf{r}_{2y}.
$$

These coefficients generally differ from those obtained when both groups are included simultaneously.

Using the group-2-only estimate, the coefficient formula can be written as

$$
\boxed{
\hat{\boldsymbol{\beta}}_1^s
=
\mathbf{R}_{1\mid2}^{-1}
\left(
\mathbf{r}_{1y}
-
\mathbf{R}_{12}
\tilde{\boldsymbol{\beta}}_2^s
\right).
}
$$

This expression shows two ways in which group 2 affects the coefficients of group 1:

1. through the Schur complement $\mathbf{R}_{1\mid2}$, which depends on the relationships between the two groups, and
2. through the term $\mathbf{R}_{12}\tilde{\boldsymbol{\beta}}_2^s$, which adjusts the correlation between group 1 and the response by the contribution associated with group 2.

```{proposition, prp-mlr-uncorrelated-groups, name="Uncorrelated Groups"}
Suppose that the two groups of explanatory variables are uncorrelated, so that

$$
\mathbf{R}_{12}=\mathbf{0}.
$$

Then

$$
\boxed{
\hat{\boldsymbol{\beta}}_1^s
=
\tilde{\boldsymbol{\beta}}_1^s
}
$$

and, analogously,

$$
\boxed{
\hat{\boldsymbol{\beta}}_2^s
=
\tilde{\boldsymbol{\beta}}_2^s.
}
$$

Thus, when the two groups are uncorrelated, including one group does not change the estimated coefficients of the other.
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-uncorrelated-groups}
If

$$
\mathbf{R}_{12}
=
\mathbf{0},
$$

then, by symmetry,

$$
\mathbf{R}_{21}
=
\mathbf{0}.
$$

Therefore,

\begin{align*}
\mathbf{R}_{1\mid2}
&=
\mathbf{R}_{11}
-
\mathbf{R}_{12}
\mathbf{R}_{22}^{-1}
\mathbf{R}_{21}
\\
&=
\mathbf{R}_{11}.
\end{align*}

The full-model coefficient becomes

\begin{align*}
\hat{\boldsymbol{\beta}}_1^s
&=
\mathbf{R}_{1\mid2}^{-1}
\left(
\mathbf{r}_{1y}
-
\mathbf{R}_{12}
\tilde{\boldsymbol{\beta}}_2^s
\right)
\\
&=
\mathbf{R}_{11}^{-1}\mathbf{r}_{1y}
\\
&=
\tilde{\boldsymbol{\beta}}_1^s.
\end{align*}

The argument for group 2 is identical.
```

</details>

Another important consequence is that zero pairwise correlation with the response does not necessarily imply a zero multiple-regression coefficient. If

$$
\mathbf{r}_{1y}=\mathbf{0},
$$

then

$$
\hat{\boldsymbol{\beta}}_1^s
=
-
\mathbf{R}_{1\mid2}^{-1}
\mathbf{R}_{12}
\tilde{\boldsymbol{\beta}}_2^s,
$$

which is not necessarily zero.

Thus, the relationship of one group with the response cannot generally be understood independently of its relationships with the other explanatory variables.

***

### Single Variable Cross-Effects

Consider the special case in which group 1 consists of a single standardized explanatory variable $\mathbf{x}_1^s$, while $\mathbf{X}_2^s$ contains all remaining explanatory variables.

Since $\mathbf{x}_1^s$ is standardized,

$$
R_{11}=1.
$$

The Schur complement is therefore the scalar

$$
R_{1\mid2}
=
1
-
\mathbf{r}_{1,2}
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2,1},
$$

where

$$
\mathbf{r}_{1,2}
=
\frac{1}{n-1}
(\mathbf{x}_1^s)'\mathbf{X}_2^s
$$

and

$$
\mathbf{r}_{2,1}
=
\mathbf{r}_{1,2}'.
$$

To understand this quantity, consider the auxiliary regression

$$
\mathbf{x}_1^s
=
\mathbf{X}_2^s\boldsymbol{\gamma}_2
+
\mathbf{e}.
$$

Its least-squares coefficient estimate is

$$
\hat{\boldsymbol{\gamma}}_2
=
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2,1}.
$$

Let $R_1^2$ denote the coefficient of determination from this regression. It measures how much of the variability in $\mathbf{x}_1^s$ can be accounted for by the remaining explanatory variables.

```{proposition, prp-mlr-schur-r2, name="Schur Complement and the Auxiliary R-Squared"}
For the auxiliary regression of $\mathbf{x}_1^s$ on $\mathbf{X}_2^s$,

$$
\boxed{
R_{1\mid2}
=
1-R_1^2.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-schur-r2}
Since $\mathbf{x}_1^s$ is standardized,

$$
\frac{1}{n-1}
(\mathbf{x}_1^s)'\mathbf{x}_1^s
=
1.
$$

The fitted values from the auxiliary regression are

$$
\hat{\mathbf{x}}_1^s
=
\mathbf{X}_2^s
\hat{\boldsymbol{\gamma}}_2.
$$

Therefore,

\begin{align*}
R_1^2
&=
\frac{
(\hat{\mathbf{x}}_1^s)'
\hat{\mathbf{x}}_1^s
}{
(\mathbf{x}_1^s)'
\mathbf{x}_1^s
}
\\
&=
\frac{
\hat{\boldsymbol{\gamma}}_2'
(\mathbf{X}_2^s)'
\mathbf{X}_2^s
\hat{\boldsymbol{\gamma}}_2
}{
n-1
}
\\
&=
\hat{\boldsymbol{\gamma}}_2'
\mathbf{R}_{22}
\hat{\boldsymbol{\gamma}}_2.
\end{align*}

Using

$$
\hat{\boldsymbol{\gamma}}_2
=
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2,1},
$$

we obtain

\begin{align*}
R_1^2
&=
\mathbf{r}_{1,2}
\mathbf{R}_{22}^{-1}
\mathbf{R}_{22}
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2,1}
\\
&=
\mathbf{r}_{1,2}
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2,1}.
\end{align*}

Consequently,

\begin{align*}
R_{1\mid2}
&=
1-
\mathbf{r}_{1,2}
\mathbf{R}_{22}^{-1}
\mathbf{r}_{2,1}
\\
&=
1-R_1^2.
\end{align*}
```

</details>

This quantity leads naturally to the Variance Inflation Factor.

```{definition, def-mlr-vif, name="Variance Inflation Factor"}
For explanatory variable $\mathbf{x}_1^s$, define

$$
\boxed{
VIF_1
=
\frac{1}{1-R_1^2},
}
$$

where $R_1^2$ is obtained by regressing $\mathbf{x}_1^s$ on all the remaining explanatory variables.
```

The name **Variance Inflation Factor** will become important when we study the variance of the estimated regression coefficients. At this stage, we interpret $VIF_1$ algebraically as a measure of how strongly $\mathbf{x}_1^s$ can be linearly represented by the remaining explanatory variables.

If

$$
R_1^2=0,
$$

then

$$
VIF_1=1.
$$

The other explanatory variables do not linearly explain $\mathbf{x}_1^s$.

As

$$
R_1^2\rightarrow1,
$$

we have

$$
\boxed{
VIF_1\rightarrow\infty.
}
$$

Thus, a large VIF indicates that little variation in $\mathbf{x}_1^s$ remains after accounting for the other explanatory variables. In the limiting case $R_1^2=1$, $\mathbf{x}_1^s$ is an exact linear combination of the remaining explanatory variables, the complete design matrix loses full column rank, and the usual unique least-squares coefficient vector no longer exists.

***

### Single Variable Coefficient Formula

For the single-variable case, the coefficient formula becomes particularly informative.

Since the coefficient from the simple regression of standardized $\mathbf{y}_s$ on standardized $\mathbf{x}_1^s$ is their sample correlation,

$$
\tilde\beta_1^s
=
r_{1y},
$$

we obtain the following result.

```{proposition, prp-mlr-single-variable-cross-effect, name="Single Variable Cross-Effect Formula"}
The coefficient of $\mathbf{x}_1^s$ in the full multiple regression satisfies

$$
\boxed{
\hat\beta_1^s
=
VIF_1
\left(
\tilde\beta_1^s
-
\mathbf{r}_{1,2}
\tilde{\boldsymbol{\beta}}_2^s
\right).
}
$$

Equivalently,

$$
\boxed{
\hat\beta_1^s
=
\frac{
r_{1y}
-
\mathbf{r}_{1,2}
\tilde{\boldsymbol{\beta}}_2^s
}{
1-R_1^2
}.
}
$$
```

<details><summary><span style="color:green">Proof</span></summary>

```{proof, pro-mlr-single-variable-cross-effect}
From the group coefficient formula,

$$
\hat\beta_1^s
=
R_{1\mid2}^{-1}
\left(
r_{1y}
-
\mathbf{r}_{1,2}
\tilde{\boldsymbol{\beta}}_2^s
\right).
$$

From the previous proposition,

$$
R_{1\mid2}
=
1-R_1^2.
$$

Therefore,

\begin{align*}
\hat\beta_1^s
&=
\frac{1}{1-R_1^2}
\left(
r_{1y}
-
\mathbf{r}_{1,2}
\tilde{\boldsymbol{\beta}}_2^s
\right)
\\
&=
VIF_1
\left(
\tilde\beta_1^s
-
\mathbf{r}_{1,2}
\tilde{\boldsymbol{\beta}}_2^s
\right).
\end{align*}
```

</details>

The formula separates two different effects created by the remaining explanatory variables:

$$
\hat\beta_1^s
=
\underbrace{VIF_1}_{\text{relationship of }x_1\text{ with the other predictors}}
\left(
\underbrace{\tilde\beta_1^s}_{\text{individual relationship with }y}
-
\underbrace{
\mathbf{r}_{1,2}
\tilde{\boldsymbol{\beta}}_2^s
}_{\text{cross-effect from the other predictors}}
\right).
$$

A large $VIF_1$ alone does not determine whether $\hat\beta_1^s$ will be large or small, because the term inside parentheses also depends on the correlations with the other explanatory variables and their relationships with the response.

***

### Generalization to Any Explanatory Variable

Variable $1$ has no special role. The same construction can be applied to any explanatory variable $\mathbf{x}_k^s$.

Let $\mathbf{X}_{-k}^s$ denote the matrix containing all explanatory variables except $\mathbf{x}_k^s$, and let $R_k^2$ be the coefficient of determination from the auxiliary regression

$$
\mathbf{x}_k^s
=
\mathbf{X}_{-k}^s
\boldsymbol{\gamma}_{-k}
+
\mathbf{e}.
$$

Then

$$
\boxed{
VIF_k
=
\frac{1}{1-R_k^2}.
}
$$

If $\tilde{\boldsymbol{\beta}}_{-k}^s$ denotes the coefficients obtained by regressing $\mathbf{y}_s$ on $\mathbf{X}_{-k}^s$, then

$$
\boxed{
\hat\beta_k^s
=
VIF_k
\left(
r_{ky}
-
\mathbf{r}_{k,-k}
\tilde{\boldsymbol{\beta}}_{-k}^s
\right).
}
$$

Thus, the coefficient of any explanatory variable in a multiple regression depends not only on its individual relationship with the response, but also on how that explanatory variable relates to all the other variables in the model.

***

#### Single Variable Cross-Effects Example

We can illustrate these quantities using the standardized GDP data from the previous section.

The theoretical notation uses $\mathbf{X}_s$ and $\mathbf{y}_s$. In the code below, these objects are temporarily renamed `X` and `y` only to simplify the R commands.

```{r gdp-cross-effects}
# Re-names Standardized Variables
y <- ys
X <- Xs

# Coefficients for the Full Model
b <- lm(y ~ X -1)$coefficients

# Individual Coefficients
# Number of Variables
P <- ncol(X)
# Vector to Save Individual Coefficients
bi <- numeric(length = P)
for(k in 1:P){
  bi[k] <- lm(y ~ X[, k] - 1)$coefficients
}

# Coefficients without variable k
bt <- matrix(data = NA, nrow = P-1, ncol = P)
for(k in 1:P){
  bt[, k] <- lm(y ~ X[, -k] - 1)$coefficients
}

# Coefficient of Determination of X_k ~ X_{-k} models and coefficients
R2k <- numeric(length = P)
for(k in 1:P){
  outReg <- lm(X[, k] ~ X[, -k] - 1)
  R2k[k] <- summary(outReg)$r.squared
}

# Cross product of correlation and coefficients
rbt <- numeric(length = P)
rXX <- cor(X)
for(k in 1:P){
  rbt[k] <- rXX[k, -k] %*% bt[, k]
}

# Formats and Show Info Table
tab           <- cbind(b, bi, R2k, 1/(1-R2k), rbt)
colnames(tab) <- c("b_full", "b_ind", "R2k", "VIF", "rb_ind")
rownames(tab) <- colnames(dat)[2:6]
knitr::kable(tab)
```

The columns of the resulting table correspond directly to the quantities developed above:

- `b_full` is $\hat\beta_k^s$, the coefficient from the full multiple regression.
- `b_ind` is $\tilde\beta_k^s=r_{ky}$, the coefficient from the individual simple regression.
- `R2k` is $R_k^2$, obtained by regressing explanatory variable $k$ on all the remaining explanatory variables.
- `VIF` is

  $$
  \frac{1}{1-R_k^2}.
  $$

- `rb_ind` is the cross-effect

  $$
  \mathbf{r}_{k,-k}
  \tilde{\boldsymbol{\beta}}_{-k}^s.
  $$

Together, these quantities illustrate the identity

$$
\hat\beta_k^s
=
VIF_k
\left(
\tilde\beta_k^s
-
\mathbf{r}_{k,-k}
\tilde{\boldsymbol{\beta}}_{-k}^s
\right).
$$

The example therefore shows explicitly why an explanatory variable can have a substantially different coefficient in a multiple regression than in a simple regression. The difference depends on both its relationships with the other explanatory variables and the relationships of those variables with the response.

***






























































## Outliers and Leverage

It is harder to identify outliers visually in multiple linear regression, specially
if the number of variables is big. The effect of an observation $i$ on the 
linear regression fit will depend on 2 things:

* The size of the residual for observation $i$, that is $\hat{e}_i^2$.
* How "close" is observation $i$ to other observations.

**Note**: For this section, we are going to use $X_i\in\dR^{1 \times p}$ as the $i-th$ row of the
design matrix. That is the $i$-th observation of the independent varaibles.

### Leverage

In multiple linear regression, **leverage** refers to the influence that a 
particular data point has on the estimation of the regression model. 
Specifically, it measures how far an observation's values for the independent 
variables deviate from the mean of those variables. Leverage identifies data 
points that could potentially have a significant effect on the regression line,
particularly those that are outliers in terms of the input variables.

Leverage values are calculated from the **hat matrix** $\bH$, which is derived
from the design matrix $\bX$ used in the regression. The diagonal elements of 
the hat matrix, $h_{ii}$, represent the leverage of each data point. 
These values range from 0 to 1, where:

- A leverage value close to 0 indicates that the point is not influential in
determining the regression line.
- A leverage value closer to 1 suggests that the point has a higher influence on
the fitted model.

High-leverage points are those whose independent variable values are far from 
the average, and they may disproportionately affect the slope and intercept of 
the regression line. These points can distort the regression model, leading to 
biased estimates if not handled properly.

A high-leverage point may not necessarily be an outlier in the response variable
(dependent variable), but it could have a significant impact on the overall 
model fit, making leverage a critical diagnostic in regression analysis.

#### Properties of Leverages

Lets analyze the structure of the hat matrix.

First, let us see that $\bH$ is symmetric semi-positive definite. Pick $z \in \dR^n$.
Recall, that $\bX' \bX$ is positive definite, then $$(\bX'\bX)^{-1}$$ is positive
definite. Then

\begin{align*}
z'Hz 
   &= \bz'\bX(\bX'\bX)^{-1}\bX'z \\
   &= (\bX'\bz)'(\bX'\bX)^{-1}(\bX'\bz) \\
   &= \bw' (\bX'\bX)^{-1} \bw & \bw = \bX'\bz \\
   &\geq 0
\end{align*}

Therefore, $\bH$ is semi-positive definite. This implies that all the eigenvalues
of $\bH$ are non-negative, that is positive or $0$.

Next, let us see that the diagonal elements are between 0 and 1.

To see this, consider leverage $i$ ($h_{ii}$), and the basis vectors $v_j \in \dR^n$ such
that are full of zeros except for entry $j$. Then:

$$h_{ij} = v_i '\bH v_j$$
in particular, leverage $i$ is given by:

$$h_{ii} = v_i '\bH v_i$$
Now, consider $\bH v_i$ and notice that:

\begin{align*}
||v_i - \bH v_i||_2^2 
  &= (v_i - \bH v_i)'(v_i - \bH v_i) \\
  &= v_i'v_i - v_i'\bH v_i - v_i'\bH v_i + v_i' \bH \bH v_i \\
  &= v_i'v_i - v_i'\bH\bH v_i - v_i'\bH\bH v_i + v_i' \bH \bH v_i \\
  &= v_i'v_i - v_i'\bH\bH v_i \\
  &= ||v_i||_2^2 - ||\bH v_i||_2^2 \\
  &= 1 - ||\bH v_i||_2^2
\end{align*}

Then

\begin{align*}
||v_i - \bH v_i||_2^2 = 1 - ||\bH v_i||_2^2
  & \implies ||v_i - \bH v_i||_2^2 + ||\bH v_i||_2^2 = 1 \\
  & \implies ||\bH v_i||_2^2 \leq 1 \\
  & \implies v_i' \bH \bH v_i  \leq 1 \\
  & \implies v_i' \bH v_i  \leq 1 \\
  & \implies h_{ii}  \leq 1 \\
\end{align*}

Since $h_{ii}=||\bH v_i||_2^2$ is a norm then $h_{ii} \geq 0$, then:

$$ 0 \leq h_{ii} \leq 1 $$
So, leverages are bounded.

Furthermore, we have that:

$$ \sum_{i=1}^p h_{ii} = p $$

To see this, we will need the following properties:

1. $\text{rank}(\mathbf{H}) = p.$
2. All the eigenvalues of $\bH$ are either 0 or 1.

As usual, we assume that $\text{rank}(\bX) = p$ and $n>p$, that is there are more observations
than independent variables. Then

\begin{align*}
\text{rank}(\bX) = p 
  &\implies \text{rank}(\bX'\bX) = p \\
  &\implies \text{rank}\left((\bX'\bX)^{-1}\right) = p \\
  &\implies \text{rank}\left(\bX(\bX'\bX)^{-1}\bX'\right) = p \\
  &\implies \text{rank}\left(\bH\right) = p \\
\end{align*}

Now, let $\lambda$ be an eigenvalue of $\bH$, with respective $\bv$ as eigenvector.
Then:

\begin{align*}
\bH \bv = \lambda \bv \\
  &\implies \bH \bH \bv = \bH \lambda \bv = \lambda \bH \bv \\
  &\implies \bH \bv = \lambda \lambda \bv = \lambda^2 \bv\\
  &\implies \lambda = 1 \land \lambda = 0
\end{align*}

Now since, $\bH$ is of rank $p$, then it must have $p$ non-zero eigenvalues. And,
since all non-zero eigenvalues are 1, then we have that:

$\sum_{i=1}^n \lambda_i = p$

that is the sum of the eigenvalues is $p$. We also know that for any square
matrix, the sum of the eigenvalues is equal to the trace of the matrix. Then:

$$ \text{trace}(\bH) = p \implies \sum_{i=1}^n h_{ii} = p$$
This is telling us important information about the leverages. You can compute a
proportional leverage as:

$$ \frac{h_{ii}}{p}$$
the proportion of the leverage $i$ out of "all" the leverage. Or in terms of 
"observations", you can compute:

$$\frac{n h_{ii}}{p}$$
if this number is several times bigger than $1$, it will be an observation with
considerable leverage.

Another way to characterize the leverages can be found as follows:

Recall that the hat matrix is given by:

$$\bH = \bX(\bX'\bX)^{-1}\bX' $$
and we can express the design matrix as follows:

$$ \bX = \left[\begin{matrix}
    X_1    \\
    X_2    \\
    \vdots \\
    X_n
  \end{matrix}\right] $$
  
and

$$ \bX' = [X_1' X_2' \ldots X_n'] $$

Then, the hat matrix is given by:

\begin{align*}
\bH 
  &= \bX(\bX'\bX)^{-1}\bX'                                                                 \\
  &= \bX(\bX'\bX)^{-1} [X_1' X_2' \ldots X_n']                                             \\
  &= \bX [(\bX'\bX)^{-1}X_1' (\bX'\bX)^{-1}X_2' \ldots (\bX'\bX)^{-1}X_n']                 \\
  &= \left[\begin{matrix}
      X_1    \\
      X_2    \\
      \vdots \\
      X_n
     \end{matrix}\right] [(\bX'\bX)^{-1}X_1' (\bX'\bX)^{-1}X_2' \ldots (\bX'\bX)^{-1}X_n'] \\
  &= \left[\begin{matrix}
      X_1 (\bX'\bX)^{-1}X_1' & X_1 (\bX'\bX)^{-1}X_2' & \ldots & X_1 (\bX'\bX)^{-1}X_n'    \\
      X_2 (\bX'\bX)^{-1}X_1' & X_2 (\bX'\bX)^{-1}X_2' & \ldots & X_2 (\bX'\bX)^{-1}X_n'    \\
      \vdots                 & \vdots                 & \vdots                             \\
      X_n (\bX'\bX)^{-1}X_1' & X_n (\bX'\bX)^{-1}X_2' & \ldots & X_n (\bX'\bX)^{-1}X_n'    \\
     \end{matrix}\right]
\end{align*}

we are going to use this expression latter, but we can also notice that:

$$h_{ij} = X_i (\bX'\bX)^{-1}X_j'$$
So we can characterize the leverages as:

$$h_{ij} = X_i (\bX'\bX)^{-1}X_j'$$

### Influential Observations

Identifying outliers in multiple regression is hard. One way to measure the influence
of an observation is to see how different are the predicted values if an observation
is removed. We will call:

* $\bX_{(i)}$ the design matrix, when the $i$-th observation is removed.
* $\by_{(i)}$ the dependent variables without the observation $i$.
* $\hat{\by}_{(i)}$ the predicted values when the $i$-th observation is removed.
* $\bH_{(i)}$ the hat matrix when the $i$-th observation is removed.

Then a measure of influence can be:

$$||\by - \hat{\by}_{(i)}||^2_2 = (\by - \hat{\by}_{(i)})'(\by - \hat{\by}_{(i)}) = \sum_{j=1}^n (\by_j - \hat{\by}_{(i),j})^2$$
Notice that:

$$\hat{\by} = \bH \by$$
and, similarly

$$\hat{\by}_{(i)} = \bH_{(i)} \by_{(i)}$$

Now, let us compute:

\begin{align*}
\hat{\by} 
  &= \bH \by \\
  &= \left[\begin{matrix}
      X_1 (\bX'\bX)^{-1}X_1' & X_1 (\bX'\bX)^{-1}X_2' & \ldots & X_1 (\bX'\bX)^{-1}X_n'    \\
      X_2 (\bX'\bX)^{-1}X_1' & X_2 (\bX'\bX)^{-1}X_2' & \ldots & X_2 (\bX'\bX)^{-1}X_n'    \\
      \vdots                 & \vdots                 & \vdots                             \\
      X_n (\bX'\bX)^{-1}X_1' & X_n (\bX'\bX)^{-1}X_2' & \ldots & X_n (\bX'\bX)^{-1}X_n'    \\
     \end{matrix}\right] \left[\begin{matrix}
      y_1    \\
      y_2    \\
      \vdots \\
      y_n
      \end{matrix}\right] \\
  &=
      \left[\begin{matrix}
        \sum_{k=1}^n X_1 (\bX'\bX)^{-1}X_k'y_k  \\
        \sum_{k=1}^n X_2 (\bX'\bX)^{-1}X_k'y_k    \\
        \vdots \\
        \sum_{k=1}^n X_n (\bX'\bX)^{-1}X_k'y_k
      \end{matrix}\right]
\end{align*}

$$
$$

Then

$$\hat{y}_j = \sum_{k=1}^n X_j (\bX'\bX)^{-1}X_k'y_k = X_j (\bX'\bX)^{-1} \sum_{k=1}^n X_k'y_k$$
in a similar way, observation :

$$\hat{y}_{(i)j} = \sum_{k=1}^n X_j (\bX_{(i)}'\bX_{(i)})^{-1}X_k'y_k = X_j (\bX_{(i)}'\bX_{(i)})^{-1} \sum_{k=1}^n X_k'y_k$$
Notice that it is not necessary to use the subscript $(i)$ when we are dealing
with single observations.

Now, we will relate $(\bX'\bX)^{-1}$ to $(\bX_{(i)}'\bX_{(i)})^{-1}$ using the 
[Sherman–Morrison formula](#sherman-morrison-formula). First notice that:

$$\bX' \bX = \bX_{(i)}'\bX_{(i)} + X_i'X_i$$

then

$$\bX_{(i)}'\bX_{(i)} = \bX' \bX - X_i'X_i$$

Now we apply the [Sherman–Morrison formula](#sherman-morrison-formula).

\begin{align*}
\left(\bX_{(i)}'\bX_{(i)}\right)^{-1} = \left(\bX' \bX - X_i'X_i\right)^{-1}
  &= \left(\bX' \bX + (-X_i)'X_i\right)^{-1} \\
  &= \left(\bX' \bX\right)^{-1} - \frac{\left(\bX' \bX\right)^{-1}(-X_i)'X_i\left(\bX' \bX\right)^{-1}}{1 + (-X_i)'\left(\bX' \bX\right)^{-1}X_i} \\
  &= \left(\bX' \bX\right)^{-1} + \frac{\left(\bX' \bX\right)^{-1}X_i'X_i\left(\bX' \bX\right)^{-1}}{1 - X_i'\left(\bX' \bX\right)^{-1}X_i} \\
  &= \left(\bX' \bX\right)^{-1} + \frac{\left(\bX' \bX\right)^{-1}X_i'X_i\left(\bX' \bX\right)^{-1}}{1 - h_{ii}}
\end{align*}

Then,

$$\hat{y}_{(i)j} = X_j \left(\bX_{(i)}'\bX_{(i)}\right)^{-1} \sum_{k \neq i}X_k'y_k = X_j\left(\bX' \bX\right)^{-1}\sum_{k \neq i}X_k'y_k + \frac{X_j\left(\bX' \bX\right)^{-1}X_i'X_i\left(\bX' \bX\right)^{-1}\sum_{k \neq i}X_k'y_k}{1 - h_{ii}}$$
Now

\begin{align*}
X_j\left(\bX' \bX\right)^{-1}\sum_{k \neq i}X_k'y_k 
  &= X_j\left(\bX' \bX\right)^{-1}\left(\sum_{k \neq i}X_k'y_k + X_i'y_i - X_i'y_i\right) \\
  &= X_j\left(\bX' \bX\right)^{-1}\left(\sum_{k = 1}^nX_k'y_k - X_i'y_i\right) \\
  &= X_j\left(\bX' \bX\right)^{-1}\sum_{k = 1}^nX_k'y_k - X_j\left(\bX' \bX\right)^{-1}X_i'y_i \\
  &= \hat{y}_j - h_{ij}y_i
\end{align*}

and

\begin{align*}
X_j\left(\bX' \bX\right)^{-1}X_i'X_i\left(\bX' \bX\right)^{-1}\sum_{k \neq i}X_k'y_k 
  &= h_{ij}X_i\left(\bX' \bX\right)^{-1}\sum_{k \neq i}X_k'y_k \\
  &= h_{ij}\sum_{k \neq i}X_i\left(\bX' \bX\right)^{-1}X_k'y_k \\
  &= h_{ij}\sum_{k \neq i}h_{ik}y_k
\end{align*}

Then

\begin{align*}
\hat{y}_{(i)j} = \hat{y}_j - h_{ij}y_i + \frac{h_{ij}\sum_{k \neq i}h_{ik}y_k}{1 - h_{ii}}
  &= \hat{y}_j - h_{ij}\left(y_i  - \frac{\sum_{k \neq i}h_{ik}y_k}{1 - h_{ii}} \right) \\
  &= \hat{y}_j - h_{ij}\left(\frac{y_i - y_ih_{ii}}{1 - h_{ii}} - \frac{\sum_{k \neq i}h_{ik}y_k}{1 - h_{ii}} \right) \\
  &= \hat{y}_j - h_{ij}\left(\frac{y_i - y_ih_{ii} - \sum_{k \neq i}h_{ik}y_k}{1 - h_{ii}} \right) \\
  &= \hat{y}_j - h_{ij}\left(\frac{y_i - \sum_{k=1}^n h_{ik}y_k}{1 - h_{ii}} \right) \\
  &= \hat{y}_j - h_{ij}\left(\frac{y_i - \hat{y}_i}{1 - h_{ii}} \right) \\
  &= \hat{y}_j - h_{ij}\left(\frac{\hat{e}_i}{1 - h_{ii}} \right)
\end{align*}

Then

\begin{align*}
\hat{y}_{(i)j} - \hat{y}_j = - h_{ij}\left(\frac{\hat{e}_i}{1 - h_{ii}} \right)
  &\implies \left(\hat{y}_{(i)j} - \hat{y}_j \right)^2= \left(h_{ij}\left(\frac{\hat{e}_i}{1 - h_{ii}} \right) \right)^2 = \frac{\hat{e}_i^2}{(1 - h_{ii})^2}h_{ij}^2 \\
  &\implies \left(\hat{\by}_{(i)} - \hat{\by} \right)'\left(\hat{\by}_{(i)} - \hat{\by} \right) = \sum_{j=1}^n \frac{\hat{e}_i^2}{(1 - h_{ii})^2}h_{ij}^2 \\
  &\implies ||\hat{\by}_{(i)} - \hat{\by}||_2^2 = \frac{\hat{e}_i^2}{(1 - h_{ii})^2} \sum_{j=1}^n h_{ij}^2 
\end{align*}

Now for any symmetric and idempotent matrix $\bM \in \dR^{n \times n}$ we have that:

$$m_{ii} = \sum_{j = 1}^n m_{ij}^2$$
Then

$$||\hat{\by}_{(i)} - \hat{\by}||_2^2 = \frac{\hat{e}_i^2}{(1 - h_{ii})^2} h_{ii}^2 = \frac{h_{ii}}{(1 - h_{ii})^2}\hat{e}_i^2 $$
That is the change in the predictions is the result of a function of the leverages
and a function of the error of observation $i$.
  
We can see this with our GDP dataset. We generate new observations, in which

```{r gdp-leverage-outliers}
# Reads Data
dat <- read.csv(file = "Gdp data.csv")

# Extracts the Data
y <- as.numeric(dat[,  1])
X <- as.matrix(dat[, -1])
n <- nrow(X)
X <- cbind(rep(1, n), X)

# Obtains the Estimates
b <- solve(t(X) %*% X, t(X) %*% y)

# Predicted Values
yh <- X %*% b

# Errors
eh <- abs(y - yh)

# Obtains the Minimum, Median and Maximum Errors
err <- quantile(eh, probs = c(0, 0.5, 1))

# Generates New Observations
# An observation at one extreme of the Observed Independent Variables
X1 <- apply(X = X, MARGIN = 2, FUN = max)
# An observation at the center of the Observed Independent Variables
X2 <- apply(X = X, MARGIN = 2, FUN = mean)
# Dependent Variables with different Error levels
y1 <- c(t(X1) %*% b) + c(0, err, err[3] * 2)
y2 <- c(t(X2) %*% b) + c(0, err, err[3] * 2)

# Outlier Level
m <- length(y1)

# Predictions without the Observations
yhi <- yh
bi  <- b
yi  <- y
Xi  <- X

# Table with Changes
tab <- matrix(data = NA, nrow = 0, ncol = 4)

for(i in 1:m){
  # Complete Data Set
  X <- rbind(Xi, X2)
  y <- c(yi, y2[i])
  
  # Coefficients
  b <- solve(t(X) %*% X, t(X) %*% y)
  
  # Predictions
  yh  <- c(X %*% b)
  yhi <- c(rbind(Xi, X2) %*% bi)
  
  # Hat matrix
  H <- X %*% solve(t(X) %*% X, t(X))
  
  # Leverage New Observation
  h <- H[n+1, n+1]
  
  # Change in Predictions
  preCha <- sum((yh - yhi)^2)
  
  # Leverage Component
  levCom <- h / ((1 - h)^2)
  
  # Error Component
  errCom <- (y[n+1] - yh[n+1])^2
  
  # Saves Values
  tab <- rbind(tab, c(preCha, levCom, errCom, levCom * errCom))
}

for(i in 1:m){
  # Complete Data Set
  X <- rbind(Xi, X1)
  y <- c(yi, y1[i])
  
  # Coefficients
  b <- solve(t(X) %*% X, t(X) %*% y)
  
  # Predictions
  yh  <- c(X %*% b)
  yhi <- c(rbind(Xi, X1) %*% bi)
  
  # Hat matrix
  H <- X %*% solve(t(X) %*% X, t(X))
  
  # Leverage New Observation
  h <- H[n+1, n+1]
  
  # Change in Predictions
  preCha <- sum((yh - yhi)^2)
  
  # Leverage Component
  levCom <- h / ((1 - h)^2)
  
  # Error Component
  errCom <- (y[n+1] - yh[n+1])^2
  
  # Saves Values
  tab <- rbind(tab, c(preCha, levCom, errCom, levCom * errCom))
}

  tab <- data.frame(tab)
  tab <- cbind(c(rep("Low", 5), rep("High", 5)), tab)
  tab <- cbind(rep(c("Minimum", "Low", "Medium", "High", "Very High"), 2), tab)
  colnames(tab) <- c("Error Level", "Leverage Level", "Sum of Sq. Diff.", "Lev. Comp.", "Err. Comp.", "Using Formula")
  knitr::kable(tab, digits = 5)
```

## Stability of the Solution

The results of the regression can be subject to numerical problems or the impact 
of changes in the data. In general, the results would depend on the stability 
of $\bX'\bX$. One way to measure the worst case scenario is the condition number
of the matrix $\bX' \bX$, denoted $\kappa(\bX' \bX)$, and is equal to:

$$ \kappa(\bX' \bX) = \frac{\lambda_{max}}{\lambda_{min}} $$
where, $\lambda_{max}$ and $\lambda_{min}$ are the maximum and minimum eigenvalues
of $\bX' \bX$.

There are some considerations, about the condition number:

- The condition number is a measure of the worst case scenario, a bigger condition
number of a matrix, does not necessarily will be more unstable than a matrix with
a lower condition number.
- $\lambda_{max} = \max_i \sigma_i^2$ and $\lambda_{max} = \min_i \sigma_i^2$, where
$\sigma_i$ are the singular values of $\bX$.

The condition number will be affected by:

1. Disparities in the magnitude of the variables.
2. The level of multicoliniarity.
3. Influential observations.

Since, the condition number is affected by the magnitude of the variables, it is common
to work with centered and standardized variables when inspecting this issues.

Another approach is to check the determinant of a matrix. A matrix would be in general
more unstable the close is it's determinant closer to 0. Indeed if the determinant is
0, then $\bX'\bX$ will not be invertible and the solution
$$ \hat{\bgb} = \left(\bX'\bX\right)^{-1}\bX' \by$$
is not applicable. However, just by looking at the determinant is hard to know
what is close to 0. Consider for example the determinant of a multiple of the identity,
then:

$\det(c\bI_n) = c^n \bI$

if $c<1$ this will make the determinant closer to zero. However the matrix $c \bI$
has condition number of $1$.

It is better to evaluate the change in the determinant after some normalization
to avoid this scenarios.

Consider, for a moment, the standardized case, with estimated coefficients:

$$ \hat{\bgb}=r_{XX}^{-1}r_{Xy}$$
then we can see how the determinant the correlation matrix, changes when we remove
a variable. Without loss of generality, we can assume we remove variable 1. Then, 
the correlation matrix will be given by:

$$r_{XX} = \left[ \begin{matrix}
r_{X_2X_2} & r_{X_2x_{1}} \\
r_{x_{1}X} & r_{x_{1}x_{1}} \\
\end{matrix} \right]$$

Then, using the [properties of the determinant](#properties-of-the-determinant)
of a block matrix, and the results of the 
[Single Variable Cross-Effects](#single-variable-cross-effects) subsection,
we have that:

\begin{align*}
\det(r_{XX}) 
  &= \det \left(
    \left[ \begin{matrix}
    r_{X_2X_2} & r_{X_2x_{1}} \\
    r_{x_{1}X} & r_{x_{1}x_{1}} \\
    \end{matrix} \right]\right) \\
  &= \det(r_{X_2X_2})\det(r_{x_{1}x_{1}} - r_{X_2x_{1}}r_{X_2X_2}^{-1}r_{x_{1}X_2}) \\
  &= \det(r_{X_2X_2})\det(r_{x_{1}|X_2}) \\
  &= \det(r_{X_2X_2})\det(1-R^2_{1}) \\
  &= \det(r_{X_2X_2})(1-R^2_{1})
\end{align*}

We can see that the determinant is closer to 0 the more correlated is the first 
variable to the rest of the variables, that is:

$$R^2_1 \to 1 \implies \det(r_{XX}) \to 0$$

In a similar way, assume that we are working with centered values, we can analyze
what would happen if we remove an observation $X_{i}$. Here we analyze $\bX' \bX$.
Again, using [properties of the determinant](#properties-of-the-determinant), and results of the 
[Outliers Identification](#outliers-identification) we have that:

$$\det(\bX_{(i)}'\bX_{(i)}) = \det(\bX_{(i)}'\bX_{(i)}) \\
  =\det(\bX'\bX' - X_iX_i') \\
  =\det(\bX'\bX')\det(1 - X_i(\bX'\bX)^{-1}X_i') \\
  =\det(\bX'\bX')\det(1 - h_{ii}) \\
  =\det(\bX'\bX')(1 - h_{ii}) \\$$

So, the determinant deteriorates the more leverage observation $i$ has.

We can see examples of the previous operations with our GDP data.

```{r condition-number-standarized}
# Read Data
dat <- read.csv("Gdp data.csv")

# Design Matrix Independent Variables
X <- as.matrix(dat[, -1])

# Centering
# Vector of Ones
v1 <- rep(1, n)
# Centering Matrix
C  <- diag(n) - (1/n) * v1 %*% t(v1)
# Independent Variables Centered
Xc <- C %*% X

# Standardizing matrix of X
DX <- diag(1/sqrt(diag(SXX)))
# Standardize X
Xs <- Xc %*% DX

# Condition numbers

print(paste0("Condition number of X'X = ",   round(kappa(t(X) %*% X), 2)))
print(paste0("Condition number of Xc'Xc = ", round(kappa(t(Xc) %*% Xc), 2)))
print(paste0("Condition number of Xs'Xs = ", round(kappa(t(Xs) %*% Xs), 2)))
```
as we can see the condition number improves as with the centering and the standardization.
However, this is not necessarily so.

Now we evaluate the effect of each variable on the determinant.

```{r condition-number-correlation}
P      <- ncol(X)
vecDet <- numeric(length = P)
vecR2  <- numeric(length = P)

# Computes
for(i in 1:P){
  vecDet[i] <- det(cor(X[, -i]))
  vecR2[i]  <- summary(lm(Xs[, i] ~ Xs[, -i] - 1))$r.squared
}

# Formats and Show Info Table
tab           <- cbind(rep(det(cor(X)), P), vecDet, vecR2)
colnames(tab) <- c("r_{X'X}", "r_{X_(i)'X_(i)}", "R2")
rownames(tab) <- colnames(dat)[2:6]
knitr::kable(tab)
```

Finally we can evaluate the effects of a new observations depending on the leverage:

```{r condition-number-new-variable}
# Read Data
dat <- read.csv("Gdp data.csv")

# Design Matrix Independent Variables
X <- as.matrix(dat[, -1])
P <- ncol(X)
n <- nrow(X)

# Centering
# Vector of Ones
v1 <- rep(1, n)
# Centering Matrix
C  <- diag(n) - (1/n) * v1 %*% t(v1)
# Independent Variables Centered
Xc <- C %*% X

# Standardizing matrix of X
DX <- diag(1/sqrt(diag(SXX)))
# Standardize X
Xs <- Xc %*% DX

# New Observations
# Very Leveraged Observation
Xn <- t(rep(3, P))
# Mildly Leveraged Observation
Xn <- rbind(Xn, t(rep(1, P)))
# No Leverage
Xn <- rbind(Xn, t(rep(0, P)))

# Number of New Observations
m <- nrow(Xn)

# Computes the Effects of the New Observation
vecDet <- numeric(length = m)
vecLev <- numeric(length = m)
for(i in 1:m){
  Z         <- rbind(Xs, Xn[i, ])
  vecDet[i] <- det(t(Z) %*% Z)
  vecLev[i] <- t(Xn[i, ]) %*% solve(t(Z) %*% Z, Xn[i, ])
}

# Formats and Show Info Table
tab           <- cbind(rep(det(t(Xs) %*% Xs), 3), vecDet, vecLev, vecLev * n / P)
colnames(tab) <- c("det(Xs_(i)'Xs_(i))","det(Xs'Xs)", "Leverage", "Obs. Equivalent")
rownames(tab) <- c("Big Leverage", "Medium Leverage", "No Leverage")
knitr::kable(tab)
```

Here’s your subsection rewritten in a **textbook style** while keeping all equations and notation unchanged. The tone and structure have been refined for clarity, precision, and smooth exposition — suitable for inclusion in a formal regression analysis chapter.

---

## Other Approaches to Regression

So far, we have focused on the **Least Squares Estimation (LSE)** framework for linear regression. While LSE provides an elegant analytical solution with well-understood statistical properties, it is by no means the only way to formulate a regression problem.

Alternative formulations arise by changing the **objective function**—the criterion used to determine the “best” coefficients. Each of these approaches yields estimators with distinct properties, often differing substantially from those of the least squares estimator.

In this section, we explore some of these alternatives. Throughout, we will denote by:

* $\hat{\boldsymbol{\beta}}$ — the least squares estimator (LSE) of the coefficients,
* $\hat{\mathbf{y}}$ — the LSE predictions for the observed data, and
* $R^2$ — the coefficient of determination, a common measure of model fit.

Each alternative approach will yield its own versions of these quantities, possessing different interpretations and statistical behaviors.

---

### Least Absolute Error Regression

Instead of minimizing the sum of squared residuals, we can minimize the **sum of absolute residuals**. This leads to the **Least Absolute Error (LAE)** or **Least Absolute Deviations (LAD)** regression problem:

$$
\min_{\boldsymbol{\beta}} \sum_{i=1}^n |y_i - \mathbf{x}_i' \boldsymbol{\beta}|
$$

Unlike the least squares estimator, the LAE estimator does not have a closed-form analytical solution. Instead, it must be computed using **numerical optimization methods**.

We denote this estimator by $\hat{\boldsymbol{\beta}}_A$.

An appealing property of LAE regression is its **robustness to outliers**. Because it penalizes large deviations linearly rather than quadratically, it tends to be less sensitive to extreme observations than the LSE.

---

### Ridge Regression

Another important modification to the least squares criterion involves adding a **penalty on the size of the coefficients**. This approach, known as **Ridge Regression**, encourages more **parsimonious** (i.e., smaller and smoother) coefficient estimates. The optimization problem is:

$$
\min_{\boldsymbol{\beta}} \sum_{i=1}^n (y_i - \mathbf{x}_i'\boldsymbol{\beta})^2 + \lambda \sum_{j=1}^n \beta_j^2
= (\mathbf{y} - \mathbf{X}\boldsymbol{\beta})'(\mathbf{y} - \mathbf{X}\boldsymbol{\beta}) + \lambda \boldsymbol{\beta}'\boldsymbol{\beta}
$$

This problem admits a closed-form solution given by:

$$
\hat{\boldsymbol{\beta}}_R = \left(\mathbf{X}'\mathbf{X} + \lambda \mathbf{I} \right)^{-1}\mathbf{X}'\mathbf{y}
$$

A notable feature of ridge regression is that it **always produces a unique solution**, even when the design matrix $\mathbf{X}$ is not of full rank, i.e.

$$
\text{rank}(\mathbf{X}) < p
$$

Ridge regression thus addresses issues of **multicollinearity** and **ill-conditioning** in $\mathbf{X}$, albeit at the cost of introducing some bias (defined in the next chapters) in the estimates.

---

### Lasso Regression

The **Lasso (Least Absolute Shrinkage and Selection Operator)** combines regularization and variable selection in a single framework. It modifies the least squares criterion by penalizing the **sum of the absolute values** of the coefficients:

$$
\min_{\boldsymbol{\beta}} \sum_{i=1}^n (y_i - \mathbf{x}_i'\boldsymbol{\beta})^2 + \lambda \sum_{j=1}^n |\beta_j|
= (\mathbf{y} - \mathbf{X}\boldsymbol{\beta})'(\mathbf{y} - \mathbf{X}\boldsymbol{\beta}) + \lambda ||\boldsymbol{\beta}||_1
$$

where $||\cdot||_1$ denotes the $\ell^1$-norm.

Unlike Ridge Regression, this problem **does not have an analytical solution** and must be solved numerically, typically through iterative algorithms such as **coordinate descent**.

The Lasso estimator, denoted $\hat{\boldsymbol{\beta}}_L$, has the remarkable property that it can **set some coefficients exactly to zero**, thereby performing **automatic feature selection**. However, its statistical properties differ significantly from both LSE and Ridge estimators.